New idtracker.ai: rethinking multi-animal tracking as a representation learning problem to increase accuracy and reduce tracking times

Version 1

Abstract

Abstract

idTracker and idtracker.ai approach multi-animal tracking from video as an image classification problem. For this classification, both rely on segments of video where all animals are visible to extract images and their identity labels. When these segments are too short, tracking can become slow and inaccurate and, if they are absent, tracking is impossible. Here, we introduce a new idtracker.ai that reframes multi-animal tracking as a representation learning problem rather than a classification task. Specifically, we apply contrastive learning to image pairs that, based on video structure, are known to belong to the same or different identities. This approach maps animal images into a representation space where they cluster by animal identity. As a result, the new idtracker.ai eliminates the need for video segments with all animals visible, is more accurate, and tracks up to 440 times faster.

Video-tracking systems that attempt to follow individuals frame-by-frame can fail during occlusions, resulting in identity swaps that accumulate over time Branson et al. (2009); Plum (2024); Chen et al. (2023); Chiara and Kim (2023); Liu et al. (2023); Bernardes et al. (2021). idTracker Pérez-Escudero et al. (2014) introduced the paradigm of animal tracking by identification from the animal images. This approach, unfeasible for humans, avoids the accumulation of errors by identity swaps during occlusions. Its successor, idtracker.ai Romero-Ferrero et al. (2019), built on this paradigm by incorporating deep learning and achieved accuracies often exceeding 99.9% in videos of up to 100 animals.

While both idTracker and idtracker.ai perform well in high-quality video, they share a limitation that can be critical in videos of lower quality or with many occlusions. To understand this limitation, consider the schematics of a video in Figure 1a. The first step of both idTracker and idtracker.ai consists in detecting instances when animals touch or cross paths (Figure 1a, shown as boxes with dashed borders and containing images of overlapping fish in this example). The video is then divided into individual fragments, each consisting of the set of images of a single individual between two animal crossings (Figure 1a shows 14 of them as rectangles with a gray background). A global fragment for a video with N animals is a collection of N fragments that coexist in one or more consecutive frames in the video (Figure 1a, the 5 fragments with blue borders are a global fragment). The significance of a global fragment is that it provides a set of images and identity labels for all the animals in the video.

Tracking by identification using deep contrastive learning.

a Schematic representation of a video with five fish. It shows 7 portions of video with animals crossing or touching (dashed-border boxes), and 14 individual fragments, sequences of images of a single individual between two crossings (gray-background boxes). The blue-border fragments form a global fragment, as there are as many individual fragments as animals and all the individual fragments coexist in one or more frames. Some pairs of images of the same animal identity are highlighted with green borders (positive images) and some images of different identities are highlighted with red borders (negative images). b A ResNet18 network with 8 outputs generates a representation of each animal image as a point in an 8-dimensional space (here shown in 2D for visualization). Each pair of images corresponds to two points in this space, separated by a Euclidean distance. The ResNet18 network is trained to minimize this distance for positive pairs and maximize it for negative pairs. c 2D t-SNE visualizations of the learned 8-dimensional representation space. Each dot represents an image of an animal from the video. As training progresses, clusters corresponding to individual animals become clearer, plotted at training with 0, 2,000, 4,000 and 15,000 batches. The t-SNE plot at 15,000 training batches is also shown color-coded by human-validated ground-truth identities. The pink rectangle at 2,000 batches of training highlights clear clusters and the orange square fuzzy clusters. d The silhouette score measures cluster coherence and increases during training, as illustrated for a video with 60 zebrafish. e A silhouette score of 0.91 corresponds to a human-validated error rate of less than 1% per image.

Figure 1—figure supplement 1. Models comparison

Figure 1—figure supplement 2. Embedding dimensions comparison

Figure 1—figure supplement 3. Dneg over Dpos comparison

Figure 1—figure supplement 4. Batch size comparison

Figure 1—figure supplement 5. Exploration and exploitation comparison

The core idea of idTracker and the original idtracker.ai is to use global fragments for the classification of images of animals into identities. In idtracker.ai, this process starts by training a convolutional neural network (CNN) with the images and labels of the global fragment that contains the longest fragment for the animal that moves the least. Once trained, the network assigns identities to all animal images in the remaining global fragments. Only global fragments meeting strict quality criteria, such as ensuring all animals in a global fragment have unique identities, are retained for further training. This iterative process of training, assigning, and selecting continues until most of the video have images assigned to identities. A second algorithm then tracks animals during crossings given that animals are already identified outside crossings.

Figure 2a (blue line) shows the accuracies of the original idtracker.ai (version 4 of the software) for a benchmark of 33 videos of zebrafish, flies and mice. These accuracies were computed using all the images of animals in the videos excluding animal crossings. Figure 2—figure Supplement 1a shows the same results but for the complete trajectory with animal crossings. The names of the videos start with a letter for the species (z,f,m), followed by the number of animals in the video, and possibly an extra number to distinguish the video if there are several of the same species and animal group size. The videos in this figure are ordered by decreasing accuracy of the original idtracker.ai results for ease of visualization. The first 15 videos are videos of zebrafish, flies and mice with an accuracy of > 99.9%. The accuracy in the remaining videos gradually decreases to 92.67% in video m_4_2, and a value of 50.4% outside the figure for video d_100_3.

Performance for a benchmark of 33 videos of flies, zebrafish and mice.

a. Median accuracy was computed using all images of animals in the videos excluding animal crossings. b. Median tracking times are shown for the scale of hours and, in the inset, for the scale of days. Supplementary Table 1, Supplementary Table 2, Supplementary Table 3 give more complete statistics (median, mean and 20-80 percentiles) for the original idtracker.ai (version 4 of the software), optimized v4 (version 5) and new idtracker.ai (version 6), respectively.

Figure 2b (blue line) shows the times that the original idtracker.ai takes to track each of the videos in the benchmark. The videos are ordered by increasing tracking times for ease of visualization. The original idtracker.ai has a faster protocol, “Protocol 2”, which works well for the simplest videos and its tracking times ranging from a few minutes to several hours. However, for complex videos, the software may switch from “Protocol 2” to “Protocol 3”, with Protocol 3 a two-step process. In the first step, all the global fragments are used to train the CNN filters. The second step proceeds like Protocol 2 but with the initial weights of the CNN filters obtained from the first step. While effective, this approach can be extremely slow, often requiring several days or weeks for a single video. Since it is stochastic whether a video is tracked using Protocol 2 or 3 (Figure 2—figure Supplement 2), a reasonable strategy to use the original idtracker.ai is to track each video multiple times until Protocol 2 successfully tracks the entire video or, when a patience threshold is reached (here set to 5 attempts), switch to Protocol 3. The tracking times shown in Figure 2b (blue line) correspond to this procedure, with the time being the accumulated time of the multiple attempts made by the software until final tracking. Some of the videos take a few minutes to track, others a few hours, and six videos take more than three days, one nearly two weeks. If we were to run idtracker.ai a single time instead of following this protocol, the tracking times for some of the videos would be longer.

We first optimized idtracker.ai by improving data loading protocols and redefining the main objects in the software (animal images and fragments) and their properties (see Methods for details). This version of the optimized original idtracker.ai (version 5 of the software) achieved better accuracies, Figure 2a (orange line), and Figure 2—figure Supplement 1a (orange line) for accuracies including animal crossings. The mean accuracy across the benchmark for this optimized version is 99.58% and 99.40% including or not animal crossings, respectively, while for the original idtracker.ai are 97.52% and 97.38%.

Even if this version also uses Protocols 2 and 3, we obtain much shorter tracking times, never longer than a day Figure 2b (orange line). On average, tracking is 13.6 times faster than with the original idtracker.ai and, for the more difficult videos, 118.4 times faster. However, waiting a day to track some videos can make a tracking pipeline too slow. To further improve accuracy and tracking times, we retained these optimizations while also changing the main logic of idtracker.ai. In the original idtracker.ai, when global fragments are short, the quality of the initial CNN is low, leading to either reduced accuracy or the triggering of the very slow Protocol 3. The new system had to be able to track without global fragments.

We reformulate multi-animal tracking as a representation learning problem. In representation learning, we learn a transformation of the input data that makes it easier to perform downstream tasks Xing et al. (2002); Bengio et al. (2013); Ericsson et al. (2022), in our case clustering into animal identities without needing identity labels. This is possible due to the structure of the video, Figure 1a. Note that pairs of images of the same individual can be obtained from the same fragment (Figure 1a, green boxes). Also, pairs of images from different individuals can be obtained from different fragments that coexist in time for one or more frames (Figure 1a, red boxes). These pairs can be used as “positive” and “negative” pairs of images for contrastive learning, a self-supervised learning framework designed to learn a representation space in which “positive” examples are close together, and “negative” examples are far apart Schroff et al. (2015); Dong and Shen (2018); KAYA and BILGE (2019); Chen et al. (2020a); Chen et al. (2020b); Guo et al. (2020); Wang et al. (2020); Yang et al. (2020).

We first evaluated neural networks suitable for contrastive learning with animal images. In addition to our previous CNN from idtracker.ai, we tested 23 networks from 8 different families of state-of-the-art convolutional neural network architectures, selected for their compatibility with consumer-grade GPUs and ability to handle small input images (20 × 20 to 100 × 100 pixels) typical in collective animal behavior videos. Among these architectures, ResNet18 He et al. (2016) was the fastest to obtain low errors (Figure 1—figure Supplement 1).

A ResNet18 with M outputs maps each input image to a point in an M-dimensional representation space (illustrated in Figure 1b as a point on a plane). Experiments showed that using M = 8 achieved faster convergence to low error (Figure 1—figure Supplement 2). ResNet18 is trained using a contrastive loss function (Chopra et al. (2005), see Methods for details). Each image in a positive or negative pair is input separately into the network, producing a point in the 8-dimensional representation space. For an image pair, we then obtain two points in an 8-dimensional space, separated by some (Euclidean) distance. The loss function is used to minimize (or maximize) this Euclidean distance for positive (or negative) pairs until the distance Dpos (or Dneg). The effect of Dpos is to prevent the collapse to a single of the positive images coming from the same fragment, allowing for a small region of the 8-dimensional representation space for all the positive pairs of the same identity. The effect of Dneg is to prevent excessive scatter of the points representing images from negative pairs. We empirically determined that Dneg/Dpos = 10 results in a faster method to obtain low error (Figure 1—figure Supplement 3), and we use Dpos = 1 and Dneg = 10.

As the model trains, the representation space becomes increasingly structured, with similar data points forming coherent clusters. Figure 1c visualizes this progression using 2D t-SNE van der Maaten and Hinton (2008) plots of the 8-dimensional representation space. After 2, 000 training batches, initial clusters emerge, and by 15,000 batches, distinct clusters corresponding to individual animals are evident. Ground truth identities verified by humans confirm that each cluster corresponds to an animal identity (Figure 1c, colored clusters).

The method to select positive and negative pairs is critical for fast learning Awasthi et al. (2022); Khosla et al. (2021); Rösch et al. (2024). This is because not all image pairs contribute equally to training. Figure 1c shows at 2, 000 training batches that some clusters well-defined (e.g. those inside the orange square) while others remain fuzzy (e.g. those inside the pink rectangle). Images in well-defined clusters have negligible impact on the loss or weight updates, as positive pairs are already close and negative pairs are sufficiently separated. Sampling from these well-defined clusters, therefore, wastes time. In contrast, fuzzy clusters contain images that still contribute significantly to the loss and benefit from further training. To address this, we developed a sampling method that prioritizes pairs from underperforming clusters requiring additional learning, while maintaining baseline sampling for all clusters based on fragment size (Methods). This ensures consistent updates across the representation space and prevents forgetting in well-defined clusters.

To assign identities to animal images, we perform K-means clustering Sculley (2010) on the points representing all images of the video in the learned 8-dimensional representation space. Each image is then assigned to a cluster with a probability that increases the closer it is to the cluster center. To evaluate clustering quality, we compute the mean Silhouette index Rousseeuw (1987), which quantifies intra-cluster cohesion and inter-cluster separation. A maximum value of 1 indicates ideal clustering. During training, the mean Silhouette index increases (Figure 1d). We empirically determined that a value of 0.91 for this index corresponds to an identity assignment error below 1% for a single image (Figure 1e). As a result, we use 0.91 as the stopping criterion for training (Methods).

After the assignment of identities to images of animals, we run some steps that are common to the previous idtracker.ai. For example, we make a final assignment of all images in fragments as each fragment must have all assignments to be the same, eliminating some errors in individual images. Also, an algorithm already present in idTracker assigns identities in the animal’s crossings taking into account that we know the identities before and after.

The new idtracker.ai has a higher accuracy than original idtracker.ai and than its optimized version, Figure 2a (magenta line). Its average accuracy in the benchmark is 99.92% and 99.78% without and with crossings, respectively, an important improvement over the original idtracker.ai (97.52% and 97.38%) and its optimized version (99.58% and 99.40%). It also gives much shorter times than the original idtracker.ai and its optimized version, Figure 2b (magenta line). It is on average 44 times faster than the original idtracker.ai and, for the more difficult videos, up to 440 times faster.

As for the original idtracker.ai, the new idtracker.ai can work well with lower resolutions, blur and video compression, and with inhomogeneous light (Figure 2—figure Supplement 4). We also compared the new idtracker.ai to TRex Walter and Couzin (2021), which is based on Protocol 2 of idtracker.ai but with additional operations like eroding crossings to make global fragments longer.

TRex gives comparable accuracies to the original idtracker.ai in the benchmark, but by avoiding Protocol 3, it is on average 31 times faster than the original idtracker.ai and up to 315 times faster (Figure 2—figure Supplement 1b). However, the new idtracker.ai is both more accurate and faster than TRex (Figure 2—figure Supplement 1). The mean accuracy of TRex across the benchmark is 98.14% and 97.89% excluding and including animal crossings, respectively. This is noticeably below the values for the new idtracker.ai of 99.92% and 99.78%, respectively. Also, the new idtracker.ai is on average 3.9 times faster and up to 16.5 times faster than TRex. Additionally, the new idtracker.ai has a memory peak lower than TRex (Figure 2—figure Supplement 3).

The new idtracker.ai also works in videos in which the original idtracker.ai does not even track because there are no global fragments. Global fragments are absent in videos with very extensive animal occlusions, for example because animals touch or cross more frequently, parts of the setup are covered, or the camera focuses on only a specific region of the setup. To study this systematically, we added a mask on the video with an angle θ (Figure 3). The tracking systems have no access to the information behind the mask. The light and dark gray regions in Figure 3 correspond to videos with no global fragments, and the original idtracker.ai and its optimized version declare tracking impossible. The new idtracker.ai, however, works well until approximately 1/4 of the setup is visible, and afterward it degrades. This also shows the limit of the new idtracker.ai. For the clustering process to be successful, we need enough coexisting individual fragments to have both positive and negative examples. Empirically, we find a deterioration with less than 0.25(N − 1) coexisting individual fragments, with N the number of animals in the video (Figure 3, dark gray region). The new idtracker.ai flags when this condition is not met.

Tracking with strong occlusions.

Accuracies when we mask a region of a video defined by an angle θ and the tracking system has no access to the information behind the mask. Light and dark gray region correspond to the angles for which no global fragments exist in the video. Dark gray regions correspond to angles for which the video does not have enough coexisting individual fragments, specifically on average less than 0.25(N − 1) coexisting individual fragments, with N the number of animals in the video. The original idtracker.ai (v4) and its optimized version (v5) cannot work in the gray regions, and new idtracker.ai is expected to deteriorate only in the dark gray region.

The final output of the new idtracker.ai consists of the x − y coordinates for each identified animal and video frame. Additionally, it provides several quality metrics: an estimate of the probability of correct identity assignment for each animal and frame, the Silhouette score as a measure of clustering quality, and the average number of coexisting individual fragments per fragment divided by (N − 1), with N the number of animal in the video, which when above 0.25(N − 1) is expected to give good results. The software can also generate a video with the computed animal trajectories for visualization, and an individual video per animal to be able to run pose estimators like the ones in Lauer et al. (2022); Pereira et al. (2022); Segalin et al. (2021); Tang et al. (2025); Biderman et al. (2024). For analysis of trajectories and spatial relationships, the user can run our Python package trajectorytools on the trajectories.

In summary, the new idtracker.ai takes an approach to tracking using representational learning to avoid the need for segments of the video in which all animals are visible. This makes the new idtracker.ai work in more videos, more accurately, much faster and with a lower memory peak.

Methods

Tested computer specifications

The software idtracker.ai depends on PyTorch and is thus compatible with any machine that can run PyTorch, including Windows, MacOS, and Linux systems. Although no specific hardware is required, a graphics card is highly recommended for hardware-accelerated machine-learning computations.

Version 6 of idtracker.ai was tested on computers running Ubuntu 24.04, Fedora 41, and Windows 11 with NVIDIA GPUs from the 1000 to the 4000 series and MacOS 15 with Metal chips. The benchmark presented in this study was performed on desktop computer running Ubuntu 24.04 LTS 64bit with a AMD Ryzen 9 5950X (32 cores at 3.4 GHz) processor, 128 GB RAM and an NVIDIA GeForce RTX 4090.

Improvements to original idtracker.ai in version 5

Following the last publication of idtracker.ai Romero-Ferrero et al. (2019), the software underwent continuous maintenance, including feature additions, performance optimizations, and hyperpa-rameter tuning (released via PyPI from March 2023 for v5.0.0 to June 2024 for v5.2.12). These updates improved the implementation and tracking pipeline but did not alter the core algorithm. Significant advancements were made in user experience, tool availability, processing speed, and memory efficiency. Below, we summarize the most notable changes.

Blob memory optimization

Blobs are defined as collections of connected pixels belonging to one or more animals. In v4, blobs stored pixel indices, causing memory usage to scale quadratically with blob size. In v5, blobs are represented by simplified contours using the Teh-Chin chain approximation Teh and Chin (1989), reducing memory usage by 93% in blob instances. This also accelerated blob-related computations (centroid, orientation, area, overlap, identification image creation, etc.).

Efficient image loading

Identification images are now efficiently loaded on demand from HDF5 files, eliminating the need to load all images into memory. This enables training with all images regardless of video length, with minimal memory usage.

Code optimization

The source code was revised to eliminate speed bottlenecks. The most impactful changes include:

Frame segmentation accelerated by 80% through optimized OpenCV usage.

Faster blob-to-blob overlap checks by first evaluating bounding boxes before deeper comparisons.

Persistent storage of blob overlap checks to avoid redundant computations when reloading data.

Efficient disk access for identification images by reading them in sorted batches, minimizing I/O overhead.

Reduced bounding box image sizes to the minimum necessary, lowering memory and processing demands.

Optimized and parallelized Torch data loaders for more efficient model training.

Caching of computationally expensive properties for blobs, fragments, and global fragments.

Sorted Fragment lists to speed up coexistence detection.

Changes to the identification protocol

In v4, identity assignments to high-confidence fragments were fixed and excluded from downstream correction, regardless of later evidence. In v5, this was relaxed for short fragments (fewer than 4 frames), allowing corrections due to their statistical unreliability and frequent image noise.

Improved graphical user interface and introduction of Exclusive ROIs

The graphical user interface was redesigned for improved usability and now includes the “Exclusive Regions of Interest” feature, which allows users to define spatially distinct regions in multi-arena experiments where animal identities are treated independently (see Figure 4 left image). It also incorporates a redesigned video generator for visualizing tracking results.

idtracker.ai new graphic user interface.

New graphics user interface (GUI) for versions v5 and v6 of idtracker.ai. On the left the segmentation GUI. On the right the Validator tool.

Validation application

A standalone GUI for inspecting and correcting tracking results. It allows users to navigate video frames, review tracked positions and metadata, detect tracking errors, and apply corrections using integrated plugins (see Figure 4, right image).

Direct integration with idmatcher.ai

A utility for matching identities across videos, originally introduced in Romero-Ferrero et al. (2023). It allows users to propagate consistent identity labels across multiple recordings, facilitating longitudinal or multi-session experiments. It is now a native feature of both v5 and v6, fully integrated into the idtracker.ai ecosystem.

Protocol details for the new idtracker.ai

In this section, we give an overview of the tracking protocol. Please refer to Appendix 1 for details.

Architectures

The contrastive learning network (Figure 1b) is a ResNet18 He et al. (2016) with a single channel in the first convolutional layer for grayscale images and 8 neurons in the last layer. The network receives grayscale images because idtracker.ai always works with grayscale converted video frames.

Loss function

The contrastive loss L for a pair of images (I, J) and label l is defined as: Here DI, J is the Euclidean distance between the embeddings of images I and J. Dpos is the maximum allowed distance between the two images of a positive pair, and Dneg, the minimum allowed distance between the two images in negative pair.

Training

ResNet18 is trained using Adam optimizer with the hyperparameters described in Kingma and Ba (2017). The learning rate is set at the value of 0.001 using training batches of 1600 images (400 positive pairs and 400 negative pairs of images). See Appendix 2 for details.

Pair selection

The selection of pairs was done by combining two sampling strategies:

Sampling fragments according to their size so that fragments containing more images are sampled more often.

Sampling fragments according to the loss function by increasing the sampling probability of pairs of fragments from whom the corresponding images had positive loss, and decreasing the sampling probability of pairs of fragments from whom the corresponding images had loss zero.

See Appendix 2 for more details on the pair sampling strategy.

Clustering and stopping criteria

For clustering, we use the minibatch K-means clustering, which significantly reduces the computation time compared to a classical implementation Sculley (2010).

Stopping of the training was done by computing the K-means clustering for a subset of (number of animals times 1,000) images, and measuring the corresponding Silhouette score (SS) Rousseeuw (1987) every number of animals times 5 batches. We stop training if there have been 30 consecutive SS evaluations without any improvement (patience of 30), or if there have been 2 consecutive SS evaluations without any improvement but the SS already achieved the target value 0.91. Check Appendix 2 for more details on the criteria to stop the training of the network.

Occlusion tests

For the occlusion tests, we took videos of freely behaving animals in a round arena (included in the benchmark) and occluded a sector of the circle between 0 and θ radians. For the tracking software, animals disappeared when entering this occluded section of the arena. The light gray area in Figure 3 corresponds to a degree of occlusion that prevents the existence of global fragments. The dark gray area in Figure 3 corresponds to a degree of occlusion where there are less than 0.25(N−1) coexisting individual fragments (N being the number of animals in the video). With these degrees of occlusion, too few animals overlap at any given time and identification is expected to deteriorate in this regime (Figure 3, dark gray region). idtracker.ai flags when this condition is not met.

Computation of tracking accuracy

Using the idtracker.ai Validator tool (see Methods), we manually generated ground-truth trajectories based on v5 outputs. This ground-truth consists on the positions and identities of all animals in each frame and their classification as either individual or crossing.

To detect tracking errors, we analyze the video frame by frame, verifying whether the predicted position of each animal deviates from the ground-truth by more than a threshold T. Errors are also recorded when the software loses the identity or fails to detect an animal in a given frame.

Tracking accuracy is then defined as one minus the proportion of errors in the trajectory. For accuracy with crossings, we consider all trajectory points, whereas for accuracy without crossings, we exclude points corresponding to crossing events in the ground-truth.

We present all results using a threshold T = 1BL with BL being a body length. We also verified that accuracy remains largely unaffected by the value of this threshold. For instance, reducing it to T = 0.5BL results in a very small change of the mean accuracy (without crossings) across the benchmark in the new idtracker.ai from 99.92% to 99.90%.

Benchmark of accuracy and tracking time

To evaluate the tracking time and accuracy of different versions of idtracker.ai and version 1.1.9 of TRex, we used a set of 33 videos with their corresponding human-validated ground-truth trajectories. Each video is 10 minutes long and features one of three species: mice, drosophila, or zebrafish, with the number of individuals ranging from 2 to 100 (see Methods).

Previous versions of idtracker.ai (v4 and v5) can resort to protocol 3 for tracking, a method that can take days to process more complex videos but is necessary when protocol 2 fails. Similarly, TRex, lacking an equivalent of protocol 3, can fail to track certain videos, leading to missing accuracy outputs (Figure 2—figure Supplement 2).

To estimate the accuracy and tracking time that a standard user might experience, we simulate a realistic user workflow. This simulation accounts for the possibility that the software may fail to track the video, prompting the user to try again with a slightly different parameter configuration, up to a certain number of attempts.

The user is given up to 5 attempts to successfully track a video. Attempts are sampled from a precomputed dataset of tracking runs. Accuracy is taken from the first successful run. The reported tracking time is the sum of the time taken by that successful run and all preceding failed attempts. In cases where all attempts fail, accuracy is determined by protocol 3 (in v4 and v5 of idtracker.ai), and tracking time includes the time required for protocol 3 plus the total time of all failed attempts. This sampling process is repeated 10,000 times per software and video to obtain statistically robust estimates of the tracking times and accuracies. Figure 2 and Figure 2—figure Supplement 1 report the median accuracies, without and with crossings, respectively, and tracking times. Supplementary Table 1, Supplementary Table 2, Supplementary Table 3, and Supplementary Table 4 present the median, mean, and the 20 and 80 percentiles in v4, v5, v6 and TRex respectively.

Dataset of tracking runs

To build the dataset of tracking runs we used for the benchmark of accuracies and times, we define input parameters through each software’s graphical interface. Fixed parameters (e.g., number of animals, regions of interest) are held constant, while those with multiple valid values are treated as variable, with their ranges annotated. In idtracker.ai, the variable parameter is the intensity_threshold, whereas in TRex, the variable parameters are threshold and track_max_speed.

Tracking is repeated for each video and software until either 5 successful runs or 35 total runs are reached. For the original version of idtracker.ai, this is limited to 3 successful runs or 7 total runs due to significantly longer tracking times. In successful runs, both accuracy and tracking time are recorded. In failed runs, when idtracker.ai defaults to protocol 3 or TRex fails to output identities (see Figure 2—figure Supplement 2), only the time until failure is recorded. For previous idtracker.ai versions (v4 and v5), failure time corresponds to the time until the software switched to protocol 3.

Each tracking run is conducted by randomly sampling values for the variable parameters from the annotated ranges and executing the full tracking process. To ensure a fair comparison, TGrabs is included when running TRex, graphical interfaces are always disabled at runtime to maximize performance, and output_interpolate_positions is enabled in TRex.

eLife Assessment

This important study introduces an advance in multi-animal tracking by reframing identity assignment as a self-supervised contrastive representation learning problem. It eliminates the need for segments of video where all animals are simultaneously visible and individually identifiable, and significantly improves tracking speed, accuracy, and robustness with respect to occlusion. This innovation has implications beyond animal tracking, potentially connecting with advances in behavioral analysis and computer vision. While the strength of support for these advances is solid overall, the presentation could be greatly improved for clarity and broader accessibility; in addition, incorporating more standard metrics in the multi-animal tracking literature would better benchmark the approach against other methods.

Reviewer #1 (Public review):

Summary:

This is a strong paper that presents a clear advance in multi-animal tracking. The authors introduce an updated version of idtracker.ai that reframes identity assignment as a contrastive learning problem rather than a classification task requiring global fragments. This change leads to gains in speed and accuracy. The method eliminates a known bottleneck in the original system, and the benchmarking across species is comprehensive and well executed. I think the results are convincing and the work is significant.

Strengths:

The main strengths are the conceptual shift from classification to representation learning, the clear performance gains, and the fact that the new version is more robust. Removing the need for global fragments makes the software more flexible in practice, and the accuracy and speed improvements are well demonstrated. The software appears thoughtfully implemented, with GUI updates and integration with pose estimators.

Weaknesses:

I don't have any major criticisms, but I have identified a few points that should be addressed to improve the clarity and accuracy of the claims made in the paper.

(1) The title begins with "New idtracker.ai," which may not age well and sounds more promotional than scientific. The strength of the work is the conceptual shift to contrastive representation learning, and it might be more helpful to emphasize that in the title rather than branding it as "new."

(2) Several technical points regarding the comparison between TRex (a system evaluated in the paper) and idtracker.ai should be addressed to ensure the evaluation is fair and readers are fully informed.

(2.1) Lines 158-160: The description of TRex as based on "Protocol 2 of idtracker.ai" overlooks several key additions in TRex, such as posture image normalization, tracklet subsampling, and the use of uniqueness feedback during training. These features are not acknowledged, and it's unclear whether TRex was properly configured - particularly regarding posture estimation, which appears to have been omitted but isn't discussed. Without knowing the actual parameters used to make comparisons, it's difficult to assess how the method was evaluated.

(2.2) Lines 162-163: The paper implies that TRex gains speed by avoiding Protocol 3, but in practice, idtracker.ai also typically avoids using Protocol 3 due to its extremely long runtime. This part of the framing feels more like a rhetorical contrast than an informative one.

(2.3) Lines 277-280: The contrastive loss function is written using the label l, but since it refers to a pair of images, it would be clearer and more precise to write it as l_{I,J}. This would help readers unfamiliar with contrastive learning understand the formulation more easily.

(2.4) Lines 333-334: The manuscript states that TRex can fail to track certain videos, but this may be inaccurate depending on how the authors classify failures. TRex may return low uniqueness scores if training does not converge well, but this isn't equivalent to tracking failure. Moreover, the metric reported by TRex is uniqueness, not accuracy. Equating the two could mislead readers. If the authors did compare outputs to human-validated data, that should be stated more explicitly.

(2.5) Lines 339-341: The evaluation approach defines a "successful run" and then sums the runtime across all attempts up to that point. If success is defined as simply producing any output, this may not reflect how experienced users actually interact with the software, where parameters are iteratively refined to improve quality.

(2.6) Lines 344-346: The simulation process involves sampling tracking parameters 10,000 times and selecting the first "successful" run. If parameter tuning is randomized rather than informed by expert knowledge, this could skew the results in favor of tools that require fewer or simpler adjustments. TRex relies on more tunable behavior, such as longer fragments improving training time, which this approach may not capture.

(2.7) Line 354 onward: TRex was evaluated using two varying parameters (threshold and track_max_speed), while idtracker.ai used only one (intensity_threshold). With a fixed number of samples, this asymmetry could bias results against TRex. In addition, users typically set these parameters based on domain knowledge rather than random exploration.

(2.8) Figure 2-figure supplement 3: The memory usage comparison lacks detail. It's unclear whether RAM or VRAM was measured, whether shared or compressed memory was included, or how memory was sampled. Since both tools dynamically adjust to system resources, the relevance of this comparison is questionable without more technical detail.

(3) While the authors cite several key papers on contrastive learning, they do not use the introduction or discussion to effectively situate their approach within related fields where similar strategies have been widely adopted. For example, contrastive embedding methods form the backbone of modern facial recognition and other image similarity systems, where the goal is to map images into a latent space that separates identities or classes through clustering. This connection would help emphasize the conceptual strength of the approach and align the work with well-established applications. Similarly, there is a growing literature on animal re-identification (ReID), which often involves learning identity-preserving representations across time or appearance changes. Referencing these bodies of work would help readers connect the proposed method with adjacent areas using similar ideas, and show that the authors are aware of and building on this wider context.

(4) Some sections of the Results text (e.g., lines 48-74) read more like extended figure captions than part of the main narrative. They include detailed explanations of figure elements, sorting procedures, and video naming conventions that may be better placed in the actual figure captions or moved to supplementary notes. Streamlining this section in the main text would improve readability and help the central ideas stand out more clearly.

Overall, though, this is a high-quality paper. The improvements to idtracker.ai are well justified and practically significant. Addressing the above comments will strengthen the work, particularly by clarifying the evaluation and comparisons.

Reviewer #2 (Public review):

This work introduces a new version of the state-of-the-art idtracker.ai software for tracking multiple unmarked animals. The authors aimed to solve a critical limitation of their previous software, which relied on the existence of "global fragments" (video segments where all animals are simultaneously visible) to train an identification classifier network, in addition to addressing concerns with runtime speed. To do this, the authors have both re-implemented the backend of their software in PyTorch (in addition to numerous other performance optimizations) as well as moving from a supervised classification framework to a self-supervised, contrastive representation learning approach that no longer requires global fragments to function. By defining positive training pairs as different images from the same fragment and negative pairs as images from any two co-existing fragments, the system cleverly takes advantage of partial (but high-confidence) tracklets to learn a powerful representation of animal identity without direct human supervision. Their formulation of contrastive learning is carefully thought out and comprises a series of empirically validated design choices that are both creative and technically sound. This methodological advance is significant and directly leads to the software's major strengths, including exceptional performance improvements in speed and accuracy and a newfound robustness to occlusion (even in severe cases where no global fragments can be detected). Benchmark comparisons show the new software is, on average, 44 times faster (up to 440 times faster on difficult videos) while also achieving higher accuracy across a range of species and group sizes. This new version of idtracker.ai is shown to consistently outperform the closely related TRex software (Walter & Couzin, 2021\), which, together with the engineering innovations and usability enhancements (e.g., outputs convenient for downstream pose estimation), positions this tool as an advancement on the state-of-the-art for multi-animal tracking, especially for collective behavior studies.

Despite these advances, we note a number of weaknesses and limitations that are not well addressed in the present version of this paper:

(1) The contrastive representation learning formulation

Contrastive representation learning using deep neural networks has long been used for problems in the multi-object tracking domain, popularized through ReID approaches like DML (Yi et al., 2014\) and DeepReID (Li et al., 2014). More recently, contrastive learning has become more popular as an approach for scalable self-supervised representation learning for open-ended vision tasks, as exemplified by approaches like SimCLR (Chen et al., 2020), SimSiam (Chen et al., 2020\), and MAE (He et al., 2021\) and instantiated in foundation models for image embedding like DINOv2 (Oquab et al., 2023). Given their prevalence, it is useful to contrast the formulation of contrastive learning described here relative to these widely adopted approaches (and why this reviewer feels it is appropriate):

(1.1) No rotations or other image augmentations are performed to generate positive examples. These are not necessary with this approach since the pairs are sampled from heuristically tracked fragments (which produces sufficient training data, though see weaknesses discussed below) and the crops are pre-aligned egocentrically (mitigating the need for rotational invariance).

(1.2) There is no projection head in the architecture, like in SimCLR. Since classification/clustering is the only task that the system is intended to solve, the more general "nuisance" image features that this architectural detail normally affords are not necessary here.

(1.3) There is no stop gradient operator like in BYOL (Grill et al., 2020\) or SimSiam. Since the heuristic tracking implicitly produces plenty of negative pairs from the fragments, there is no need to prevent representational collapse due to class asymmetry. Some care is still needed, but the authors address this well through a pair sampling strategy (discussed below).

(1.4) Euclidean distance is used as the distance metric in the loss rather than cosine similarity as in most contrastive learning works. While cosine similarity coupled with L2-normalized unit hypersphere embeddings has proven to be a successful recipe to deal with the curse of dimensionality (with the added benefit of bounded distance limits), the authors address this through a cleverly constructed loss function that essentially allows direct control over the intra- and inter-cluster distance (D\_pos and D\_neg). This is a clever formulation that aligns well with the use of K-means for the downstream assignment step.

No concerns here, just clarifications for readers who dig into the review. Referencing the above literature would enhance the presentation of the paper to align with the broader computer vision literature.

(2) Network architecture for image feature extraction backbone

As most of the computations that drive up processing time happen in the network backbone, the authors explored a variety of architectures to assess speed, accuracy, and memory requirements. They land on ResNet18 due to its empirically determined performance. While the experiments that support this choice are solid, the rationale behind the architecture selection is somewhat weak. The authors state that:

"\[W\]e tested 23 networks from 8 different families of state-of-the-art convolutional neural network architectures, selected for their compatibility with consumer-grade GPUs and ability to handle small input images (20 × 20 to 100 × 100 pixels) typical in collective animal behavior videos."

(2.1) Most modern architectures have variants that are compatible with consumer-grade GPUs. This is true of, for example, HRNet (Wang et al., 2019), ViT (Dosovitskiy et al., 2020), SwinT (Liu et al., 2021), or ConvNeXt (Liu et al., 2022), all of which report single GPU training and fast runtime speeds through lightweight configuration or subsequent variants, e.g., MobileViT (Mehta et al., 2021). The authors may consider revising that statement or providing additional support for that claim (e.g., empirical experiments) given that these have been reported to outperform ResNet18 across tasks.

(2.2) The compatibility of different architectures with small image sizes is configurable. Most convolutional architectures can be readily adapted to work with smaller image sizes, including 20x20 crops. With their default configuration, they lose feature map resolution through repeated pooling and downsampling steps, but this can be readily mitigated by swapping out standard convolutions with dilated convolutions and/or by setting the stride of pooling layers to 1, preserving feature map resolution across blocks. While these are fairly straightforward modifications (and are even compatible with using pretrained weights), an even more trivial approach is to pad and/or resize the crops to the default image size, which is likely to improve accuracy at a possibly minimal memory and runtime cost. These techniques may even improve the performance with the architectures that the authors did test out.

(2.3) The authors do not report whether the architecture experiments were done with pretrained or randomly initialized weights.

(2.4) The authors do not report some details about their ResNet18 design, specifically whether a global pooling layer is used and whether the output fully connected layer has any activation function. Additionally, they do not report the version of ResNet18 employed here, namely, whether the BatchNorm and ReLU are applied after (v1) or before (v2) the conv layers in the residual path.

(3) Pair sampling strategy

The authors devised a clever approach for sampling positive and negative pairs that is tailored to the nature of the formulation. First, since the positive and negative labels are derived from the co-existence of pretracked fragments, selection has to be done at the level of fragments rather than individual images. This would not be the case if one of the newer approaches for contrastive learning were employed, but it serves as a strength here (assuming that fragment generation/first pass heuristic tracking is achievable and reliable in the dataset). Second, a clever weighted sampling scheme assigns sampling weights to the fragments that are designed to balance "exploration and exploitation". They weigh samples both by fragment length and by the loss associated with that fragment to bias towards different and more difficult examples.

(3.1) The formulation described here resembles and uses elements of online hard example mining (Shrivastava et al., 2016), hard negative sampling (Robinson et al., 2020\), and curriculum learning more broadly. The authors may consider referencing this literature (particularly Robinson et al., 2020\) for inspiration and to inform the interpretation of the current empirical results on positive/negative balancing.

(4) Speed and accuracy improvements

The authors report considerable improvements in speed and accuracy of the new idTracker (v6) over the original idTracker (v4?) and TRex. It's a bit unclear, however, which of these are attributable to the engineering optimizations (v5?) versus the representation learning formulation.

(4.1) Why is there an improvement in accuracy in idTracker v5 (L77-81)? This is described as a port to PyTorch and improvements largely related to the memory and data loading efficiency. This is particularly notable given that the progression went from 97.52% (v4; original) to 99.58% (v5; engineering enhancements) to 99.92% (v6; representation learning), i.e., most of the new improvement in accuracy owes to the "optimizations" which are not the central emphasis of the systematic evaluations reported in this paper.

(4.2) What about the speed improvements? Relative to the original (v4), the authors report average speed-ups of 13.6x in v5 and 44x in v6. Presumably, the drastic speed-up in v6 comes from a lower Protocol 2 failure rate, but v6 is not evaluated in Figure 2 - figure supplement 2.

(5) Robustness to occlusion

A major innovation enabled by the contrastive representation learning approach is the ability to tolerate the absence of a global fragment (contiguous frames where all animals are visible) by requiring only co-existing pairs of fragments owing to the paired sampling formulation. While this removes a major limitation of the previous versions of idtracker.ai, its evaluation could be strengthened. The authors describe an ablation experiment where an arc of the arena is masked out to assess the accuracy under artificially difficult conditions. They find that the v6 works robustly up to significant proportions of occlusions, even when doing so eliminates global fragments.

(5.1) The experiment setup needs to be more carefully described.

What does the masking procedure entail? Are the pixels masked out in the original video or are detections removed after segmentation and first pass tracking is done?

What happens at the boundary of the mask? (Partial segmentation masks would throw off the centroids, and doing it after original segmentation does not realistically model the conditions of entering an occlusion area.)

Are fragments still linked for animals that enter and then exit the mask area?

How is the evaluation done? Is it computed with or without the masked region detections?

(5.2) The circular masking is perhaps not the most appropriate for the mouse data, which is collected in a rectangular arena.

(5.3) The number of co-existing fragments, which seems to be the main determinant of performance that the authors derive from this experiment, should be reported for these experiments. In particular, a "number of co-existing fragments" vs accuracy plot would support the use of the 0.25(N-1) heuristic and would be especially informative for users seeking to optimize experimental and cage design. Additionally, the number of co-existing fragments can be artificially reduced in other ways other than a fixed occlusion, including random dropout, which would disambiguate it from potential allocentric positional confounds (particularly relevant in arenas where egocentric pose is correlated with allocentric position).

(6) Robustness to imaging conditions

The authors state that "the new idtracker.ai can work well with lower resolutions, blur and video compression, and with inhomogeneous light (Figure 2 - figure supplement 4)." (L156).

Despite this claim, there are no speed or accuracy results reported for the artificially corrupted data, only examples of these image manipulations in the supplementary figure.

(7) Robustness across longitudinal or multi-session experiments

The authors reference idmatcher.ai as a compatible tool for this use case (matching identities across sessions or long-term monitoring across chunked videos), however, no performance data is presented to support its usage.

This is relevant as the innovations described here may interact with this setting. While deep metric learning and contrastive learning for ReID were originally motivated by these types of problems (especially individuals leaving and entering the FOV), it is not clear that the current formulation is ideally suited for this use case. Namely, the design decisions described in point 1 of this review are at times at odds with the idea of learning generalizable representations owing to the feature extractor backbone (less scalable), low-dimensional embedding size (less representational capacity), and Euclidean distance metric without hypersphere embedding (possible sensitivity to drift).

It's possible that data to support point 6 can mitigate these concerns through empirical results on variations in illumination, but a stronger experiment would be to artificially split up a longer video into shorter segments and evaluate how generalizable and stable the representations learned in one segment are across contiguous ("longitudinal") or discontiguous ("multi-session") segments.

Reviewer #3 (Public review):

Summary:

The authors propose a new version of idTracker.ai for animal tracking. Specifically, they apply contrastive learning to embed cropped images of animals into a feature space where clusters correspond to individual animal identities.

Strengths:

By doing this, the new software alleviates the requirement for so-called global fragments - segments of the video, in which all entities are visible/detected at the same time - which was necessary in the previous version of the method. In general, the new method reduces the tracking time compared to the previous versions, while also increasing the average accuracy of assigning the identity labels.

Weaknesses:

The general impression of the paper is that, in its current form, it is difficult to disentangle the old from the new method and understand the method in detail. The manuscript would benefit from a major reorganization and rewriting of its parts. There are also certain concerns about the accuracy metric and reducing the computational time.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary

This is a strong paper that presents a clear advance in multi-animal tracking. The authors introduce an updated version of idtracker.ai that reframes identity assignment as a contrastive learning problem rather than a classification task requiring global fragments. This change leads to gains in speed and accuracy. The method eliminates a known bottleneck in the original system, and the benchmarking across species is comprehensive and well executed. I think the results are convincing and the work is significant.

Strengths

The main strengths are the conceptual shift from classification to representation learning, the clear performance gains, and the fact that the new version is more robust. Removing the need for global fragments makes the software more flexible in practice, and the accuracy and speed improvements are well demonstrated. The software appears thoughtfully implemented, with GUI updates and integration with pose estimators.

Weaknesses

I don't have any major criticisms, but I have identified a few points that should be addressed to improve the clarity and accuracy of the claims made in the paper.

(1) The title begins with "New idtracker.ai," which may not age well and sounds more promotional than scientific. The strength of the work is the conceptual shift to contrastive representation learning, and it might be more helpful to emphasize that in the title rather than branding it as "new."

We considered using “Contrastive idtracker.ai”. However, we thought that readers could then think that we believe they could use both the old idtracker.ai or this contrastive version. But we want to say that the new version is the one to use as it is better in both accuracy and tracking times. We think “New idtracker.ai” communicates better that this version is the version we recommend.

(2) Several technical points regarding the comparison between TRex (a system evaluated in the paper) and idtracker.ai should be addressed to ensure the evaluation is fair and readers are fully informed.

(2.1) Lines 158-160: The description of TRex as based on "Protocol 2 of idtracker.ai" overlooks several key additions in TRex, such as posture image normalization, tracklet subsampling, and the use of uniqueness feedback during training. These features are not acknowledged, and it's unclear whether TRex was properly configured - particularly regarding posture estimation, which appears to have been omitted but isn't discussed. Without knowing the actual parameters used to make comparisons, it's difficult to dassess how the method was evaluated.

We added the information about the key additions of TRex in the section “The new idtracker.ai uses representation learning”, lines 153-157. Posture estimation in TRex was not explicitly used but neither disabled during the benchmark; we clarified this in the last paragraph of “Benchmark of accuracy and tracking time”, lines 492-495.

(2.2) Lines 162-163: The paper implies that TRex gains speed by avoiding Protocol 3, but in practice, idtracker.ai also typically avoids using Protocol 3 due to its extremely long runtime. This part of the framing feels more like a rhetorical contrast than an informative one.

We removed this, see new lines 153-157.

(2.3) Lines 277-280: The contrastive loss function is written using the label l, but since it refers to a pair of images, it would be clearer and more precise to write it as l_{I,J}. This would help readers unfamiliar with contrastive learning understand the formulation more easily.

We added this change in lines 613-620.

(2.4) Lines 333-334: The manuscript states that TRex can fail to track certain videos, but this may be inaccurate depending on how the authors classify failures. TRex may return low uniqueness scores if training does not converge well, but this isn't equivalent to tracking failure. Moreover, the metric reported by TRex is uniqueness, not accuracy. Equating the two could mislead readers. If the authors did compare outputs to human-validated data, that should be stated more explicitly.

We observed TRex crashing without outputting any trajectories on some occasions (Appendix 1—figure 1), and this is what we labeled as “failure”. These failures happened in the most difficult videos of our benchmark, that’s why we treated them the same way as idtracker.ai going to P3. We clarified this in new lines 464-469.

The accuracy measured in our benchmark is not estimated but it is human-validated (see section Computation of tracking accuracy in Appendix 1). Both softwares report some quality estimators at the end of a tracking (“estimated accuracy” for idtracker.ai and "uniqueness” for TRex) but these were not used in the benchmark.

(2.5) Lines 339-341: The evaluation approach defines a "successful run" and then sums the runtime across all attempts up to that point. If success is defined as simply producing any output, this may not reflect how experienced users actually interact with the software, where parameters are iteratively refined to improve quality.

Yes, our benchmark was designed to be agnostic to the different experiences of the user. Also, our benchmark was designed for users that do not inspect the trajectories to choose parameters again not to leave room for potential subjectivity.

(2.6) Lines 344-346: The simulation process involves sampling tracking parameters 10,000 times and selecting the first "successful" run. If parameter tuning is randomized rather than informed by expert knowledge, this could skew the results in favor of tools that require fewer or simpler adjustments. TRex relies on more tunable behavior, such as longer fragments improving training time, which this approach may not capture.

We precisely used the TRex parameter track_max_speed to elongate fragments for optimal tracking. Rather than randomized parameter tuning, we defined the “valid range” for this parameter so that all values in it would produce a decent fragment structure. We used this procedure to avoid worsening those methods that use more parameters.

(2.7) Line 354 onward: TRex was evaluated using two varying parameters (threshold and track_max_speed), while idtracker.ai used only one (intensity_threshold). With a fixed number of samples, this asymmetry could bias results against TRex. In addition, users typically set these parameters based on domain knowledge rather than random exploration.

idtracker.ai and TRex have several parameters. Some of them have a single correct value (e.g. number of animals) or the default value that the system computes is already good (e.g. minimum blob size). For a second type of parameters, the system finds a value that is in general not as good, so users need to modify them. In general, users find that for this second type of parameter there is a valid interval of possible values, from which they need to choose a single value to run the system. idtracker.ai has intensity_threshold as the only parameter of this second type and TRex has two: threshold and track_max_speed. For these parameters, choosing one value or another within the valid interval can give different tracking results. Therefore, when we model a user that wants to run the system once except if it goes to P3 (idtracker.ai) or except if it crashes (TRex), it is these parameters we sample from within the valid interval to get a different value for each run of the system. We clarify this in lines 452-469 of the section “Benchmark of accuracy and tracking time”.

Note that if we chose to simply run old idtracker.ai (v4 or v5) or TRex a single time, this would benefit the new idtracker.ai (v6). This is because old idtracker.ai can enter the very slow protocol 3 and TRex can fail to track. So running old idtracker.ai or TRex up to 5 times until old idtracker.ai does not use Protocol 3 and TRex does not fail is to make them as good as they can be with respect to the new idtracker.ai.

(2.8) Figure 2-figure supplement 3: The memory usage comparison lacks detail. It's unclear whether RAM or VRAM was measured, whether shared or compressed memory was included, or how memory was sampled. Since both tools dynamically adjust to system resources, the relevance of this comparison is questionable without more technical detail.

We modified the text in the caption (new Figure 1-figure supplement 2) adding the kind of memory we measured (RAM) and how we measured it. We already have a disclaimer for this plot saying that memory management depends on the machine's available resources. We agree that this is a simple analysis of the usage of computer resources.

(3) While the authors cite several key papers on contrastive learning, they do not use the introduction or discussion to effectively situate their approach within related fields where similar strategies have been widely adopted. For example, contrastive embedding methods form the backbone of modern facial recognition and other image similarity systems, where the goal is to map images into a latent space that separates identities or classes through clustering. This connection would help emphasize the conceptual strength of the approach and align the work with well-established applications. Similarly, there is a growing literature on animal re-identification (ReID), which often involves learning identity-preserving representations across time or appearance changes. Referencing these bodies of work would help readers connect the proposed method with adjacent areas using similar ideas, and show that the authors are aware of and building on this wider context.

We have now added a new section in Appendix 3, “Differences with previous work in contrastive/metric learning” (lines 792-841) to include references to previous work and a description of what we do differently.

(4) Some sections of the Results text (e.g., lines 48-74) read more like extended figure captions than part of the main narrative. They include detailed explanations of figure elements, sorting procedures, and video naming conventions that may be better placed in the actual figure captions or moved to supplementary notes. Streamlining this section in the main text would improve readability and help the central ideas stand out more clear

Thank you for pointing this out. We have rewritten the Results, for example streamlining the old lines 48-74 (new lines 42-48) by moving the comments about names, files and order of videos to the caption of Figure 1.

Overall, though, this is a high-quality paper. The improvements to idtracker.ai are well justified and practically significant. Addressing the above comments will strengthen the work, particularly by clarifying the evaluation and comparisons.

We thank the reviewer for the detailed suggestions. We believe we have taken all of them into consideration to improve the ms.

Reviewer #2 (Public review):

Summary:

This work introduces a new version of the state-of-the-art idtracker.ai software for tracking multiple unmarked animals. The authors aimed to solve a critical limitation of their previous software, which relied on the existence of "global fragments" (video segments where all animals are simultaneously visible) to train an identification classifier network, in addition to addressing concerns with runtime speed. To do this, the authors have both re-implemented the backend of their software in PyTorch (in addition to numerous other performance optimizations) as well as moving from a supervised classification framework to a self-supervised, contrastive representation learning approach that no longer requires global fragments to function. By defining positive training pairs as different images from the same fragment and negative pairs as images from any two co-existing fragments, the system cleverly takes advantage of partial (but high-confidence) tracklets to learn a powerful representation of animal identity without direct human supervision. Their formulation of contrastive learning is carefully thought out and comprises a series of empirically validated design choices that are both creative and technically sound. This methodological advance is significant and directly leads to the software's major strengths, including exceptional performance improvements in speed and accuracy and a newfound robustness to occlusion (even in severe cases where no global fragments can be detected). Benchmark comparisons show the new software is, on average, 44 times faster (up to 440 times faster on difficult videos) while also achieving higher accuracy across a range of species and group sizes. This new version of idtracker.ai is shown to consistently outperform the closely related TRex software (Walter & Couzin, 2021\), which, together with the engineering innovations and usability enhancements (e.g., outputs convenient for downstream pose estimation), positions this tool as an advancement on the state-of-the-art for multi-animal tracking, especially for collective behavior studies.

Despite these advances, we note a number of weaknesses and limitations that are not well addressed in the present version of this paper:

Weaknesses

(1) The contrastive representation learning formulation. Contrastive representation learning using deep neural networks has long been used for problems in the multi-object tracking domain, popularized through ReID approaches like DML (Yi et al., 2014) and DeepReID (Li et al., 2014). More recently, contrastive learning has become more popular as an approach for scalable self-supervised representation learning for open-ended vision tasks, as exemplified by approaches like SimCLR (Chen et al., 2020), SimSiam (Chen et al., 2020), and MAE (He et al., 2021) and instantiated in foundation models for image embedding like DINOv2 (Oquab et al., 2023). Given their prevalence, it is useful to contrast the formulation of contrastive learning described here relative to these widely adopted approaches (and why this reviewer feels it is appropriate):

(1.1) No rotations or other image augmentations are performed to generate positive examples. These are not necessary with this approach since the pairs are sampled from heuristically tracked fragments (which produces sufficient training data, though see weaknesses discussed below) and the crops are pre-aligned egocentrically (mitigating the need for rotational invariance).

(1.2) There is no projection head in the architecture, like in SimCLR. Since classification/clustering is the only task that the system is intended to solve, the more general "nuisance" image features that this architectural detail normally affords are not necessary here.

(1.3) There is no stop gradient operator like in BYOL (Grill et al., 2020\) or SimSiam. Since the heuristic tracking implicitly produces plenty of negative pairs from the fragments, there is no need to prevent representational collapse due to class asymmetry. Some care is still needed, but the authors address this well through a pair sampling strategy (discussed below).

(1.4) Euclidean distance is used as the distance metric in the loss rather than cosine similarity as in most contrastive learning works. While cosine similarity coupled with L2-normalized unit hypersphere embeddings has proven to be a successful recipe to deal with the curse of dimensionality (with the added benefit of bounded distance limits), the authors address this through a cleverly constructed loss function that essentially allows direct control over the intra- and inter-cluster distance (D\_pos and D\_neg). This is a clever formulation that aligns well with the use of K-means for the downstream assignment step.

No concerns here, just clarifications for readers who dig into the review. Referencing the above literature would enhance the presentation of the paper to align with the broader computer vision literature.

Thank you for this detailed comparison. We have now added a new section in Appendix 3, “Differences with previous work in contrastive/metric learning” (lines 792-841) to include references to previous work and a description of what we do differently, including the points raised by the reviewer.

(2) Network architecture for image feature extraction backbone. As most of the computations that drive up processing time happen in the network backbone, the authors explored a variety of architectures to assess speed, accuracy, and memory requirements. They land on ResNet18 due to its empirically determined performance. While the experiments that support this choice are solid, the rationale behind the architecture selection is somewhat weak. The authors state that: "We tested 23 networks from 8 different families of state-of-the-art convolutional neural network architectures, selected for their compatibility with consumer-grade GPUs and ability to handle small input images (20 × 20 to 100 × 100 pixels) typical in collective animal behavior videos."

(2.1) Most modern architectures have variants that are compatible with consumer-grade GPUs. This is true of, for example, HRNet (Wang et al., 2019), ViT (Dosovitskiy et al., 2020), SwinT (Liu et al., 2021), or ConvNeXt (Liu et al., 2022), all of which report single GPU training and fast runtime speeds through lightweight configuration or subsequent variants, e.g., MobileViT (Mehta et al., 2021). The authors may consider revising that statement or providing additional support for that claim (e.g., empirical experiments) given that these have been reported to outperform ResNet18 across tasks.

Following the recommendation of the reviewer, we tested the architectures SwinT, ConvNeXt and ViT. We found out that none of them outperformed ResNet18 since they all showed a slower learning curve. This would result in higher tracking times. These tests are now included in the section “Network architecture” (lines 550-611).

(2.2) The compatibility of different architectures with small image sizes is configurable. Most convolutional architectures can be readily adapted to work with smaller image sizes, including 20x20 crops. With their default configuration, they lose feature map resolution through repeated pooling and downsampling steps, but this can be readily mitigated by swapping out standard convolutions with dilated convolutions and/or by setting the stride of pooling layers to 1, preserving feature map resolution across blocks. While these are fairly straightforward modifications (and are even compatible with using pretrained weights), an even more trivial approach is to pad and/or resize the crops to the default image size, which is likely to improve accuracy at a possibly minimal memory and runtime cost. These techniques may even improve the performance with the architectures that the authors did test out.

The only two tested architectures that require a minimum image size are AlexNet and DenseNet. DenseNet proved to underperform ResNet18 in the videos where the images are sufficiently large. We have tested AlexNet with padded images to see that it also performs worse than ResNet18 (see Appendix 3—figure 1).

We also tested the initialization of ResNet18 with pre-trained weights from ImageNet (in Appendix 3—figure 2) and it proved to bring no benefit to the training speed (added in lines 591-592).

(2.3) The authors do not report whether the architecture experiments were done with pretrained or randomly initialized weights.

We adapted the text to make it clear that the networks are always randomly initialized (lines 591-592, lines 608-609 and the captions of Appendix 3—figure 1 and 2).

(2.4) The authors do not report some details about their ResNet18 design, specifically whether a global pooling layer is used and whether the output fully connected layer has any activation function. Additionally, they do not report the version of ResNet18 employed here, namely, whether the BatchNorm and ReLU are applied after (v1) or before (v2) the conv layers in the residual path.

We use ResNet18 v1 with no activation function nor bias in its last layer (this has been clarified in the lines 606-608). Also, by design, ResNet has a global average pool right before the last fully connected layer which we did not remove. In response to the reviewer, Resnet18 v2 was tested and its performance is the same as that of v1 (see Appendix 3—figure 1 and lines 590-591).

(3) Pair sampling strategy. The authors devised a clever approach for sampling positive and negative pairs that is tailored to the nature of the formulation. First, since the positive and negative labels are derived from the co-existence of pretracked fragments, selection has to be done at the level of fragments rather than individual images. This would not be the case if one of the newer approaches for contrastive learning were employed, but it serves as a strength here (assuming that fragment generation/first pass heuristic tracking is achievable and reliable in the dataset). Second, a clever weighted sampling scheme assigns sampling weights to the fragments that are designed to balance "exploration and exploitation". They weigh samples both by fragment length and by the loss associated with that fragment to bias towards different and more difficult examples.

(3.1) The formulation described here resembles and uses elements of online hard example mining (Shrivastava et al., 2016), hard negative sampling (Robinson et al., 2020\), and curriculum learning more broadly. The authors may consider referencing this literature (particularly Robinson et al., 2020\) for inspiration and to inform the interpretation of the current empirical results on positive/negative balancing.

Following this recommendation, we added references of hard negative mining in the new section “Differences with previous work in contrastive/metric learning”, lines 792-841. Regarding curriculum learning, even though in spirit it might have parallels with our sampling method in the sense that there is a guided training of the network, we believe the approach is more similar to an exploration-exploitation paradigm.

(4) Speed and accuracy improvements. The authors report considerable improvements in speed and accuracy of the new idTracker (v6) over the original idTracker (v4?) and TRex. It's a bit unclear, however, which of these are attributable to the engineering optimizations (v5?) versus the representation learning formulation.

(4.1) Why is there an improvement in accuracy in idTracker v5 (L77-81)? This is described as a port to PyTorch and improvements largely related to the memory and data loading efficiency. This is particularly notable given that the progression went from 97.52% (v4; original) to 99.58% (v5; engineering enhancements) to 99.92% (v6; representation learning), i.e., most of the new improvement in accuracy owes to the "optimizations" which are not the central emphasis of the systematic evaluations reported in this paper.

V5 was a two year-effort designed to improve time efficiency of v4. It was also a surprise to us that accuracy was higher, but that likely comes from the fact that the substituted code from v4 contained some small bug/s. The improvements in v5 are retained in v6 (contrastive learning) and v6 has higher accuracy and shorter tracking times. The difference in v6 for this extra accuracy and shorter tracking times is contrastive learning.

(4.2) What about the speed improvements? Relative to the original (v4), the authors report average speed-ups of 13.6x in v5 and 44x in v6. Presumably, the drastic speed-up in v6 comes from a lower Protocol 2 failure rate, but v6 is not evaluated in Figure 2 - figure supplement 2.

Idtracker.ai v5 runs an optimized Protocol 2 and, sometimes, the Protocol 3. But v6 doesn’t run either of them. While P2 is still present in v6 as a fallback protocol when contrastive fails, in our v6 benchmark P2 was never needed. So the v6 speedup comes from replacing both P2 and P3 with the contrastive algorithm.

(5) Robustness to occlusion. A major innovation enabled by the contrastive representation learning approach is the ability to tolerate the absence of a global fragment (contiguous frames where all animals are visible) by requiring only co-existing pairs of fragments owing to the paired sampling formulation. While this removes a major limitation of the previous versions of idtracker.ai, its evaluation could be strengthened. The authors describe an ablation experiment where an arc of the arena is masked out to assess the accuracy under artificially difficult conditions. They find that the v6 works robustly up to significant proportions of occlusions, even when doing so eliminates global fragments.

(5.1) The experiment setup needs to be more carefully described.

(5.1.1) What does the masking procedure entail? Are the pixels masked out in the original video or are detections removed after segmentation and first pass tracking is done?

The mask is defined as a region of interest in the software. This means that it is applied at the segmentation step where the video frame is converted to a foreground-background binary image. The region of interest is applied here, converting to background all pixels not inside of it. We clarified this in the newly added section Occlusion tests, lines 240-244.

(5.1.2) What happens at the boundary of the mask? (Partial segmentation masks would throw off the centroids, and doing it after original segmentation does not realistically model the conditions of entering an occlusion area.)

Animals at the boundaries of the mask are partially detected. This can change the location of their detected centroid. That’s why, when computing the ground-truth accuracy for these videos, only the groundtruth centroids that were at minimum 15 pixels further from the mask were considered. We clarified this in the newly added section Occlusion tests, lines 248-251.

(5.1.3) Are fragments still linked for animals that enter and then exit the mask area?

No artificial fragment linking was added in these videos. Detected fragments are linked the usual way. If one animal hides into the mask, the animal disappears so the fragment breaks. We clarified this in the newly added section Occlusion tests, lines 245-247.

(5.1.4) How is the evaluation done? Is it computed with or without the masked region detections?

The groundtruth used to validate these videos contains the positions of all animals at all times. But only the positions outside the mask at each frame were considered to compute the tracking accuracy. We clarified this in the newly added section Occlusion tests, lines 248-251.

(5.2) The circular masking is perhaps not the most appropriate for the mouse data, which is collected in a rectangular arena.

We wanted to show the same proof of concept in different videos. For that reason, we used to cover the arena parametrized by an angle. In the rectangular arena the circular masking uses an external circle, so it is covering the rectangle parametrized by an angle.

(5.3) The number of co-existing fragments, which seems to be the main determinant of performance that the authors derive from this experiment, should be reported for these experiments. In particular, a "number of co-existing fragments" vs accuracy plot would support the use of the 0.25(N-1) heuristic and would be especially informative for users seeking to optimize experimental and cage design. Additionally, the number of co-existing fragments can be artificially reduced in other ways other than a fixed occlusion, including random dropout, which would disambiguate it from potential allocentric positional confounds (particularly relevant in arenas where egocentric pose is correlated with allocentric position).

We included the requested analysis about the fragment connectivity in Figure 3-figure supplement 1. We agree that there can be additional ways of reducing co-existing fragments, but we think the occlusion tests have the additional value that there are many real experiments similar to this test.

(6) Robustness to imaging conditions. The authors state that "the new idtracker.ai can work well with lower resolutions, blur and video compression, and with inhomogeneous light (Figure 2 - figure supplement 4)." (L156). Despite this claim, there are no speed or accuracy results reported for the artificially corrupted data, only examples of these image manipulations in the supplementary figure.

We added this information in the same image, new Figure 1 - figure supplement 3.

(7) Robustness across longitudinal or multi-session experiments. The authors reference idmatcher.ai as a compatible tool for this use case (matching identities across sessions or long-term monitoring across chunked videos), however, no performance data is presented to support its usage. This is relevant as the innovations described here may interact with this setting. While deep metric learning and contrastive learning for ReID were originally motivated by these types of problems (especially individuals leaving and entering the FOV), it is not clear that the current formulation is ideally suited for this use case. Namely, the design decisions described in point 1 of this review are at times at odds with the idea of learning generalizable representations owing to the feature extractor backbone (less scalable), low-dimensional embedding size (less representational capacity), and Euclidean distance metric without hypersphere embedding (possible sensitivity to drift). It's possible that data to support point 6 can mitigate these concerns through empirical results on variations in illumination, but a stronger experiment would be to artificially split up a longer video into shorter segments and evaluate how generalizable and stable the representations learned in one segment are across contiguous ("longitudinal") or discontiguous ("multi-session") segments.

We have now added a test to prove the reliability of idmatcher.ai in v6. In this test, 14 videos are taken from the benchmark and split in two non-overlapping parts (with a 200 frames gap in between). idmatcher.ai is run between the two parts presenting a 100% accuracy identity matching across all of them (see section “Validity of idmatcher.ai in the new idtracker.ai”, lines 969-1008).

We thank the reviewer for the detailed suggestions. We believe we have taken all of them into consideration to improve the ms.

Reviewer #3 (Public review):

Summary

The authors propose a new version of idTracker.ai for animal tracking. Specifically, they apply contrastive learning to embed cropped images of animals into a feature space where clusters correspond to individual animal identities.

Strengths

By doing this, the new software alleviates the requirement for so-called global fragments - segments of the video, in which all entities are visible/detected at the same time - which was necessary in the previous version of the method. In general, the new method reduces the tracking time compared to the previous versions, while also increasing the average accuracy of assigning the identity labels.

Weaknesses

The general impression of the paper is that, in its current form, it is difficult to disentangle the old from the new method and understand the method in detail. The manuscript would benefit from a major reorganization and rewriting of its parts. There are also certain concerns about the accuracy metric and reducing the computational time.

We have made the following modifications in the presentation:

(1) We have added section tiles to the main text so it is clearer what tracking system we are referring to. For example, we now have sections “Limitation of the original idtracker.ai”, “Optimizing idtracker.ai without changes in the learning method” and “The new idtracker.ai uses representation learning”.

(2) We have completely rewritten all the text of the ms until we start with contrastive learning. Old L20-89 is now L20-L66, much shorter and easier to read.

(3) We have rewritten the first 3 paragraphs in the section “The new idtracker.ai uses representation learning” (lines 68-92).

(4) We now expanded Appendix 3 to discuss the details of our approach (lines 539-897). It discusses in detail the steps of the algorithm, the network architecture, the loss function, the sampling strategy, the clustering and identity assignment, and the stopping criteria in training

(5) To cite previous work in detail and explain what we do differently, we have now added in Appendix 3 the new section “Differences with previous work in contrastive/metric learning” (lines 792-841).

Regarding accuracy metrics, we have replaced our accuracy metric with the standard metric IDF1. IDF1 is the standard metric that is applied to systems in which the goal is to maintain consistent identities across time. See also the section in Appendix 1 "Computation of tracking accuracy” (lines 414-436) explaining IDF1 and why this is an appropriate metric for our goal.

Using IDF1 we obtain slightly higher accuracies for the idtracker.ai systems. This is the comparison of mean accuracy over all our benchmark for our previous accuracy score and the new one for the full trajectories:

v4: 97.42% -> 98.24%

v5: 99.41% -> 99.49%

v6: 99.74% -> 99.82%

trex: 97.89% -> 97.89%

We thank the reviewer for the suggestions about presentation and about the use of more standard metrics.

Recommendations for the authors:

Reviewer #2 (Recommendations for the authors):

(1) Figure 1a: A graphical legend inset would make it more readable since there are multiple colors, line styles, and connecting lines to parse out.

Following this recommendation, we added a graphical legend in the old Figure 1 (new Figure 2).

(2) L46: "have images" → "has images".

We applied this correction. Line 35.

(3) L52: "videos start with a letter for the species (z,**f**,m)", but "d" is used for fly videos.

We applied this correction in the caption of Figure 1.

(4) L62: "with Protocol 3 a two-step process" → "with Protocol 3 being a two-step process".

We rewrote this paragraph without mentioning Protocol 3, lines 37-41.

(5) L82-89: This is the main statement of the problems that are being addressed here (speed and relaxing the need for global fragments). This could be moved up, emphasized, and made clearer without the long preamble and results on the engineering optimizations in v5. This lack of linearity in the narrative is also evident in the fact that after Figure 1a is cited, inline citations skip to Figure 2 before returning to Figure 1 once the contrastive learning is introduced.

We have rewritten all the text until the contrastive learning, (old lines 20-89 are now lines 20-66). The text is shorter, more linear and easier to read.

(6) L114: "pairs until the distance D_{pos}" → "pairs until the distance approximates D_{pos}".

We rewrote as “ pairs until the distance 𝐷pos (or 𝐷neg) is reached” in line 107.

(7) L570: Missing a right parenthesis in the equation.

We no longer have this equation in the ms.

(8) L705: "In order to identify fragments we, not only need" → "In order to identify fragments, we not only need".

We applied this correction, Line 775.

(9) L819: "probably distribution" → "probability distribution".

We applied this correction, Line 776.

(10) L833: "produced the best decrease the time required" → "produced the best decrease of the time required".

We applied this correction, Line 746.

Reviewer #3 (Recommendations for the authors):

(1) We recommend rewriting and restructuring the manuscript. The paper includes a detailed explanation of the previous approaches (idTracker and idTracker.ai) and their limitations. In contrast, the description of the proposed method is short and unstructured, which makes it difficult to distinguish between the old and new methods as well as to understand the proposed method in general. Here are a few examples illustrating the problem.

(1.1) Only in line 90 do the authors start to describe the work done in this manuscript. The previous 3 pages list limitations of the original method.

We have now divided the main text into sections, so it is clearer what is the previous method (“Limitation of the original idtracker.ai”, lines 28-51), the new optimization we did of this method (“Optimizing idtracker.ai without changes in the learning method”, lines 52-66) and the new contrastive approach that also includes the optimizations (“The new idtracker.ai uses representation learning”, lines 66-164). Also, the new text has now been streamlined until the contrastive section, following your suggestion. You can see that in the new writing the three sections are 25 , 15 and 99 lines. The more detailed section is the new system, the other two are needed as reference, to describe which problem we are solving and the extra new optimizations.

(1.2) The new method does not have a distinct name, and it is hard to follow which idtracker.ai is a specific part of the text referring to. Not naming the new method makes it difficult to understand.

We use the name new idtracker.ai (v6) so it becomes the current default version. v5 is now obsolete, as well as v4. And from the point of view of the end user, no new name is needed since v6 is just an evolution of the same software they have been using. Also, we added sections in the main text to clarify the ideas in there and indicate the version of idtracker.ai we are referring to.

(1.3) There are "Protocol 2" and "Protocol 3" mixed with various versions of the software scattered throughout the text, which makes it hard to follow. There should be some systematic naming of approaches and a listing of results introduced.

Following this recommendation we no longer talk about the specific protocols of the old version of idtracker.ai in the main text. We rewritten the explanation of these versions in a more clear and straightforward way, lines 29-36.

(2) To this end, the authors leave some important concepts either underexplained or only referenced indirectly via prior work. For example, the explanation of how the fragments are created (line 15) is only explained by the "video structure" and the algorithm that is responsible for resolving the identities during crossings is not detailed (see lines 46-47, 149-150). Including summaries of these elements would improve the paper's clarity and accessibility.

We listed the specific sections from our previous publication where the reader can find information about the entire tracking pipeline (lines 539-549). This way, we keep the ms clear and focused on the new identification algorithm while indicating where to find such information.

(3) Accuracy metrics are not clear. In line 319, the authors define it as based on "proportion of errors in the trajectory". This proportion is not explained. How is the error calculated if a trajectory is lost or there are identity swaps? Multi-object tracking has a range of accuracy metrics that account for such events but none of those are used by the authors. Estimating metrics that are common for MOT literature, for example, IDF1, MOTA, and MOTP, would allow for better method performance understanding and comparison.

In the new ms, we replaced our accuracy metric with the standard metric IDF1. IDF1 is the standard metric that is applied to systems in which the goal is to maintain consistent identities across time. See also the section in Appendix 1 "Computation of tracking accuracy” explaining why IDF1 and not MOTA or MOTP is the adequate metric for a system that wants to give correct tracking by identification in time. See lines 416-436.

Using IDF1 we obtain slightly higher accuracies for the idtracker.ai systems. This is the comparison of mean accuracy four our previous accuracy and the new one for the full trajectories:

v4: 97.42% -> 98.24%

v5: 99.41% -> 99.49%

v6: 99.74% -> 99.82%

trex: 97.89% -> 97.89%

(4) Additionally, the authors distinguish between tracking with and without crossings, but do not provide statistics on the frequency of crossings per video. It is also unclear how the crossings are considered for the final output. Including information such as the frame rate of the videos would help to better understand the temporal resolution and the differences between consecutive frames of the videos.

We added this information in the Appendix 1 “Benchmark of accuracy and tracking time”, lines 445-451. The framerate in our benchmark videos goes from 25 to 60 fps (average of 37 fps). On average 2.6% of the blobs are crossings (1.1% for zebrafish 0.7% for drosophila 9.4% for mice).

(5) In the description of the dataset used for evaluation (lines 349-365), the authors describe the random sampling of parameter values for each tracking run. However, it is unclear whether the same values were used across methods. Without this clarification, comparisons between the proposed method, older versions, and TRex might be biased due to lucky parameter combinations. In addition, the ranges from which the values were randomly sampled were also not described.

Only one parameter is shared between idtracker.ai and TRex: intensity_threshold (in idtracker.ai) and threshold (in TRex). Both are conceptually equivalent but differ in their numerical values since they affect different algorithms. V4, v5, and TRex each required the same process of independent expert visual inspection of the segmentation to select the valid value range. Since versions 5 and 6 use exactly the same segmentation algorithm, they share the same parameter ranges.

All the ranges of valid values used in our benchmark are public here https://drive.google.com/drive/folders/1tFxdtFUudl02ICS99vYKrZLeF28TiYpZ as stated in the section “Data availability”, lines 227-228.

(6) Lines 122-123, Figure 1c. "batches" - is an imprecise metric of training time as there is no information about the batch size.

We clarified the Figure caption, new Figure 2c.

(7) Line 145 - "we run some steps... For example..." leaves the method description somewhat unclear. It would help if you could provide more details about how the assignments are carried out and which metrics are being used.

Following this recommendation, we listed the specific sections from our previous publication where the reader can find information about the entire tracking pipeline (lines 539-549). This way, we keep the ms clear and focused on the new identification algorithm while indicating where to find such information.

(8) Figure 3. How is tracking accuracy assessed with occlusions? Are the individuals correctly recognized when they reappear from the occluded area?

The groundtruth for this video contains the positions of all animals at all times. Only the groundtruth points inside the region of interest are taken into account when computing the accuracy. When the tracking reaches high accuracy, it means that animals are successfully relabeled every time they enter the non-masked region. Note that this software works all the time by identification of animals, so crossings and occlusion are treated the same way. What is new here is that the occlusions are so large that there are no global fragments. We clarified this in the new section “Occlusion tests” in Methods, lines 239-251.

(9) Lines 185-187 this part of the sentence is not clear.

We rewrote this part in a clearer way, lines 180-182.

(10) The authors also highlight the improved runtime performance. However, they do not provide a detailed breakdown of the time spent on each component of the tracking/training pipeline. A timing breakdown would help to compare the training duration with the other components. For example, the calculation of the Silhouette Score alone can be time-consuming and could be a bottleneck in the training process. Including this information would provide a clearer picture of the overall efficiency of the method.

We measured that the training of ResNet takes on average in our benchmark 47% of the tracking time (we added this information line 551 section “Network Architecture”). In this training stage the bottleneck becomes the network forward and backward pass, limited by the GPU performance. All other processes happening during training have been deeply optimized and parallelized when needed so their contribution to the training time is minimal. Apart from the training, we also measured 24.4% of the total tracking time spent in reading and segmenting the video files and 11.1% in processing the identification images and detecting crossings.

(11) An important part of the computational cost is related to model training. It would be interesting to test whether a model trained on one video of a specific animal type (e.g., zebrafish_5) generalizes to another video of the same type (e.g., zebrafish_7). This would assess the model's generalizability across different videos of the same species and spare a lot of compute. Alternatively, instead of training a model from scratch for each video, the authors could also consider training a base model on a superset of images from different videos and then fine-tuning it with a lower learning rate for each specific video. This could potentially save time and resources while still achieving good performance.

Already before v6, there was the possibility for the user to start training the identification network by copying the final weights from another tracking session. This knowledge transfer feature is still present in v6 and it still decreases the training times significatively. This information has been added in Appendix 4, lines 906-909.

We have already begun working on the interesting idea of a general base model but it brings some complex challenges. It could be a very useful new feature for future idtracker.ai releases.

We thank the reviewer for the many suggestions. We have implemented all of them.

Version 2

Abstract

idTracker and idtracker.ai approach multi-animal tracking from video as an image classification problem. For this classification, both rely on segments of video where all animals are visible to extract images and their identity labels. When these segments are too short, tracking can become slow and inaccurate and, if they are absent, tracking is impossible. Here, we introduce a new idtracker.ai that reframes multi-animal tracking as a representation learning problem rather than a classification task. Specifically, we apply contrastive learning to image pairs that, based on video structure, are known to belong to the same or different identities. This approach maps animal images into a representation space where they cluster by animal identity. As a result, the new idtracker.ai eliminates the need for video segments with all animals visible, is more accurate, and tracks up to 700 times faster.

Main text

Video-tracking systems that attempt to follow individuals frame-by-frame can fail during oc-clusions, resulting in identity swaps that accumulate over time Branson et al. (2009); Plum (2024); Chen et al. (2023); Chiara and Kim (2023); Liu et al. (2023); Bernardes et al. (2021). idTracker Pérez-Escudero et al. (2014) introduced the paradigm of animal tracking by identification from the animal images. This approach, unfeasible for humans, avoids the accumulation of errors by identity swaps during occlusions. Its successor, idtracker.ai Romero-Ferrero et al. (2019), built on this paradigm by incorporating deep learning and achieved accuracies often exceeding 99.9% in videos of up to 100 animals.

Limitation of the original idtracker.ai

The core idea of the original idtracker.ai is to use a segment of the video in which all animals are visible to extract a set of images and identity labels for each individual. These labeled images are used to train a convolutional neural network (CNN) with one class per animal. Once trained, the network assigns identities to other segments in which all animals are visible. Only segments with identity assignments that meet strict quality criteria are retained, and their images and labels are also used for further training of the CNN. This iterative process of training, assigning, and selecting continues until most of the animal images in the video have been assigned to identities. A second algorithm then tracks animals during crossings.

If no segment exists in which all animals are visible, idtracker.ai cannot start. More commonly, such segments can be short and result in a CNN of low quality. To improve performance in this case, idtracker.ai pretrains the CNN using the entire video, but this process is slow. As a consequence, when the segments in which all animals visible are too short, accuracy might be lower and tracking time longer.

Figure 1a (blue line) gives the accuracies obtained with the original idtracker.ai in our bench-mark of 33 videos (see Appendix 1 for details of benchmarking). The first 15 videos of the bench-mark are videos of zebrafish, flies (drosophila) and mice with for which of the original idtracker.ai has accuracy of > 99.9%. In the remaining videos the accuracy decreases, reaching 93.77% in video m_4_2, and 69.66% in video d_100_3, which lies outside the plotted range. These accuracies were computed using all animal images in the video while excluding animal crossings. Figure 1—figure Supplement 1a shows the corresponding results for complete trajectories that include crossings.

Performance for a benchmark of 33 videos of flies, zebrafish and mice.

a. Median accuracy was computed using all images of animals in the videos excluding animal crossings. The videos are ordered by decreasing accuracy of the original idtracker.ai results for ease of visualization. b. Median tracking times are shown for the scale of hours and, in the inset, for the scale of days. The videos are ordered by increasing tracking times in the original idtracker.ai results for ease of visualization. Supplementary Table 1, Supplementary Table 2, Supplementary Table 3 give more complete statistics (median, mean and 20-80 percentiles) for the original idtracker.ai (version 4 of the software), optimized v4 (version 5) and new idtracker.ai (version 6), respectively. The names of the videos in a. and b. start with a letter for the species (z,d,m), followed by the number of animals in the video, and possibly an extra number to distinguish the video if there are several of the same species and animal group size.

Performance for the benchmark with full trajectories with animal crossings.

a. Median accuracy was computed using all images of animals in the videos including animal crossings. b. Median tracking times. Supplementary Table 1, Supplementary Table 2, Supplementary Table 3 and Supplementary Table 4 give more complete statistics (median, mean and 20-80 percentiles) for the original idtracker.ai (version 4 of the software), optimized v4 (version 5), new idtracker.ai (version 6) and TRex, respectively.

Robustness to blurring and light conditions.

First column: Unmodified video zebrafish_60_1. Second column: zebrafish_60_1 with a gaussian blurring of 1 pixel plus a resolution reduction to 40% of the original plus MJPG video compression. Versions 5 of idtracker.ai fails to track this videos with a 82% accuracy. Third column: Videos of 60 zebrafish with manipulated light conditions (same test as in idtracker.ai Romero-Ferrero et al. (2019)). First row: Uniform light conditions across the arena (zebrafish_60_1). Second row: Similar setup but with lights off in the bottom and right side of the arena.

Memory usage across the different softwares.

The solid line is a logarithmic fit to the (RAM) memory peak as a function of the number of blobs in a video. The Python tool psutil.virtual_memory().used was used to measure the memory usage 5 times per second during the tracking. The presented value is the maximum measured value minus the baseline measured before the tracking start. Disclaimer: Both softwares include automatic optimizations that adjust based on machine resources, so results may vary on systems with less available memory. These results were measured on computers with the specifications in Methods.

Figure 1b (blue line) shows the times that the original idtracker.ai takes to track each of the videos in the benchmark. Some of the videos take a few minutes to track, others a few hours, and six videos take more than three days, one nearly two weeks.

Optimizing idtracker.ai without changes in the learning method

We first optimized idtracker.ai without changing how we identify animals. We improved data loading and redesigned the main objects in the software (see Appendix 2 for details). This version of the optimized original idtracker.ai (version 5 of the software) achieved higher accuracies, Figure 1a (orange line), and Figure 1—figure Supplement 1a (orange line) for results that include animal crossings. The mean accuracy across the benchmark is 99.63% without crossings and 99.49% with crossings, compared with 98.39% without crossings and 98.24% with crossings for the original id-tracker.ai.

Tracking times were also significantly reduced. As shown in Figure 1b (orange line), no video took longer than a day. On average, tracking is 13.5 times faster than with the original version and 120.1 times faster for the more difficult videos. Even so, tracking times of up to one day may still be a limiting factor in some pipelines.

To further improve both accuracy and tracking speed, we retained these optimizations while changing the core identification algorithm to eliminate the need for segments of video in which all animals are visible.

The new idtracker.ai uses representation learning

We reformulate multi-animal tracking as a representation learning problem. In representation learning, we learn a transformation of the input data that makes it easier to perform downstream tasks Xing et al. (2002); Bengio et al. (2013); Ericsson et al. (2022). In our case, the downstream task is to cluster images into animal identities without requiring identity labels.

This reformulation is made possible by the inherent structure of the video, illustrated in Figure 2a. First, we detect specific moments in the video when animals touch or cross paths. These are shown in Figure 2a as boxes with dashed borders that contain images of overlapping animals. We can then divide the rest of the video into individual fragments, each consisting of the set of images of a single individual between two animal crossings. Figure 2a shows 14 such fragments as rectangles with a gray background. In addition, a video with N animals may contain global fragments, that is, a collections of N individual fragments that coexist in one or more consecutive frames. An example is shown in Figure 2a by the five fragments with blue borders. These global fragments are used by the original idtracker.ai to train a CNN. In the new approach, we do not assume that global fragments exist.

Tracking by identification using deep contrastive learning.

a Schematic representation of a video with five fish. b A ResNet18 network with 8 outputs generates a representation of each animal image as a point in an 8-dimensional space (here shown in 2D for visualization). Each pair of images corresponds to two points in this space, separated by a Euclidean distance. The ResNet18 network is trained to minimize this distance for positive pairs and maximize it for negative pairs. c 2D t-SNE visualizations of the learned 8-dimensional representation space. Each dot represents an image of an animal from the video. As training progresses, clusters corresponding to individual animals become clearer, plotted at training with 0, 2,000, 4,000 and 15,000 batches (each batch contains 400 positive and 400 negative pairs of images). The t-SNE plot at 15,000 training batches is also shown color-coded by human-validated ground-truth identities. The pink rectangle at 2,000 batches of training highlights clear clusters and the orange square fuzzy clusters. d The Silhouette score measures cluster coherence and increases during training, as illustrated for a video with 60 zebrafish. e A Silhouette score of 0.91 corresponds to a human-validated error rate of less than 1% per image.

Representation learning becomes feasible in this context because we can obtain both positive and negative image pairs without using identity labels. Positive pairs are images of the same individual obtained from within the same individual fragment (Figure 2a, green boxes). Negative pairs are images of different individuals taken from different individual fragments that coexist in time for one or more frames (Figure 2a, red boxes). We can then use the positive and negative pairs of images for contrastive learning, a self-supervised learning framework designed to learn a representation space in which positive examples are close together, and negative examples are far apart Schroff et al. (2015); Dong and Shen (2018); KAYA and BiLGE (2019); Chen et al. (2020a,b); Guo et al. (2020); Wang et al. (2020); Yang et al. (2020). This formulation is supported by the success of deep metric and contrastive learning methods (see Appendix 3 for comparison with previous work).

We first evaluated which neural networks are suitable for contrastive learning with animal images. In addition to our previous CNN from idtracker.ai, we tested 31 networks from 10 different families of state-of-the-art convolutional neural networks and transformer architectures, selected for their compatibility with consumer-grade GPUs and ability to handle small input images (20 × 20 to 100 × 100 pixels) typical in collective animal behavior videos. Among these architectures, ResNet18 (v1) He et al. (2016a) without pretrained weights was the fastest to obtain low errors (see Appendix 3).

A ResNet18 with M outputs maps each input image to a point in an M-dimensional representation space (illustrated in Figure 2b as a point on a plane). Experiments showed that using M = 8 achieved faster convergence to low error (see Appendix 3). ResNet18 is trained using a contrastive loss function (Chopra et al. (2005), see Appendix 3 for details). Each image in a positive or negative pair is input separately into the network, producing a point in the 8-dimensional representation space. For an image pair, we then obtain two points in an 8-dimensional space, separated by some (Euclidean) distance. The optimization of the loss function minimizes (or maximizes) this Euclidean distance for positive (or negative) pairs until the distance Dpos (or Dneg) is reached. The effect of Dpos is to prevent the collapse to a single of the positive images coming from the same fragment, allowing for a small region of the 8-dimensional representation space for all the positive pairs of the same identity. The effect of Dneg is to prevent excessive scatter of the points representing images from negative pairs. We empirically determined that Dneg/Dpos = 10 results in a faster method to obtain low error (see Appendix 3), and we use Dpos = 1 and Dneg = 10.

As the model trains, the representation space becomes increasingly structured, with similar data points forming coherent clusters. Figure 2c visualizes this progression using 2D t-SNE van der Maaten and Hinton (2008) plots of the 8-dimensional representation space. After 2, 000 training batches, initial clusters emerge, and by 15,000 batches, distinct clusters corresponding to individual animals are evident. Ground truth identities verified by humans confirm that each cluster corresponds to an animal identity (Figure 2c, colored clusters).

The method to select positive and negative pairs is critical for fast learning Awasthi et al. (2022); Khosla et al. (2021); Rösch et al. (2024). This is because not all image pairs contribute equally to training. Figure 2c shows at 2, 000 training batches that some clusters are well-defined (e.g. those inside the pink rectangle) while others remain fuzzy (e.g. those inside the orange square). Images in well-defined clusters have negligible impact on the loss or weight updates, as positive pairs are already close and negative pairs are sufficiently separated. Sampling from these well-defined clusters, therefore, wastes time. In contrast, fuzzy clusters contain images that still contribute significantly to the loss and benefit from further training. To address this, we developed a sampling method that prioritizes pairs from underperforming clusters requiring additional learning, while maintaining baseline sampling for all clusters based on fragment size (see Appendix 3). This ensures consistent updates across the representation space and prevents forgetting in well-defined clusters.

To assign identities to animal images, we perform K-means clustering Sculley (2010) on the points representing all images of the video in the learned 8-dimensional representation space. Each image is then assigned to a cluster with a probability that increases the closer it is to the cluster center. To evaluate clustering quality, we compute the mean Silhouette score Rousseeuw (1987), which quantifies intra-cluster cohesion and inter-cluster separation. A maximum value of 1 indicates ideal clustering. During training, the mean Silhouette score increases (Figure 2d). We empirically determined that a value of 0.91 for this index corresponds to an identity assignment error below 1% for a single image (Figure 2e). As a result, we use 0.91 as the stopping criterion for training (see Appendix 3).

After the assignment of identities to images of animals, we run some steps that are common to the previous idtracker.ai. For example, we make a final assignment of all images in fragments as each fragment must have all assignments to be the same, eliminating some errors in individual images. Also, an algorithm already present in idTracker assigns identities in the animal’s crossings taking into account that we know the identities before and after.

The new idtracker.ai (v6) has a higher accuracy than original idtracker.ai (v4) and than its optimized version (v5), Figure 1a (purple line). Its average accuracy in the benchmark is 99.92% and 99.82% without and with crossings, respectively, an important improvement over the original idtracker.ai v4 (98.39% and 98.24%) and its optimized version v5 (99.63% and 99.49%). It also gives much shorter times than the original idtracker.ai (v4) and its optimized version (v5), Figure 1b (purple line). It is on average 90 times faster than the original idtracker.ai (v4) and, for the more difficult videos, up to 712 times faster.

As for the original idtracker.ai, the new idtracker.ai can work well with lower resolutions, blur and video compression, and with inhomogeneous light (Figure 1—figure Supplement 2). We also compared the new idtracker.ai to TRex Walter and Couzin (2021), which is based on idtracker.ai without pretraining and with additional operations like eroding crossings to make global fragments longer, posture image normalization, tracklet subsampling, and the use of uniqueness feedback during training. TRex gives comparable accuracies to the original idtracker.ai in the benchmark, and it is on average 31 times faster than the original idtracker.ai and up to 316 times faster (Figure 1—figure Supplement 1b). However, the new idtracker.ai is both more accurate and faster than TRex (Figure 1—figure Supplement 1). The mean accuracy of TRex across the benchmark is 98.14% and 97.89% excluding and including animal crossings, respectively. This is noticeably below the values for the new idtracker.ai of 99.92% and 99.82%, respectively. Also, the new idtracker.ai is on average 4.5 times faster and up to 16.5 times faster than TRex. Additionally, the new idtracker.ai has a memory peak lower than TRex (Figure 1—figure Supplement 3).

No need for global fragments

The new idtracker.ai also works in videos in which the original idtracker.ai does not even track because there are no global fragments. Global fragments are absent in videos with very extensive animal occlusions, for example because animals touch or cross more frequently, parts of the setup are covered, or the camera focuses on only a specific region of the setup.

To systematically evaluate this, we digitally masked videos with a sector of angle θ (Figure 3). Masked pixels were enforced as background during segmentation, causing animals to disappear from the video when entering the occluded region. Consequently, each re-entry into the visible area requires re-identification with no knowledge of the identities in the occluded region (see Methods for more details about the occlusion tests).

Tracking with strong occlusions.

Accuracies when we mask a region of a video defined by an angle θ and the tracking system has no access to the information behind the mask. Light and dark gray region correspond to the angles for which no global fragments exist in the video. Dark gray regions correspond to angles for which the video has a fragment connectivity lower than 0.5, with the fragment connectivity defined as the average number of other fragments each fragment coexists with, divided by N − 1, with N the total number of animals; see Figure 3—figure Supplement 1, for an analysis justifying this value of 0.5. The original idtracker.ai (v4) and its optimized version (v5) cannot work in the gray regions, and new idtracker.ai is expected to deteriorate only in the dark gray region.

Fragment connectivity analysis.

The relation between fragment connectivity and the corresponding accuracy for every video and angle θ of the new id-tracker.ai in Figure 3. Fragment connectivity is calculated as the average number of other fragments each fragment coexists with, divided by N − 1, with N the total number of animals. Values below 0.5 are associated with low accuracy, and idtracker.ai alerts the user in such cases.

The light and dark gray regions in Figure 3 correspond to videos with no global fragments, and the original idtracker.ai and its optimized version declare tracking impossible in these regions. The new idtracker.ai, however, works well until approximately 1/4 of the setup is visible, and afterward it degrades. This shows the limit of the new idtracker.ai. The deterioration happens because, for the clustering process to be successful, we need enough coexisting individual fragments to have both positive and negative examples. We measure this using fragment connectivity, defined as the average number of other fragments a fragment coexists with, divided by N − 1, with N the total number of animals in the video. Empirically, we find that a fragment connectivity below 0.5 corresponds to low accuracies Figure 3—figure Supplement 1. The new idtracker.ai flags when this condition of low fragment connectivity takes place, which we indicate in Figure 3 with the dark gray regions.

Output of new idtracker.ai

The final output of the new idtracker.ai consists of the x − y coordinates for each identified animal and video frame. Additionally, it provides an estimation of the achieved accuracy, the Silhouette score as a measure of clustering quality, and the probability of correct identity assignment for each animal and frame. The new idtracker.ai also includes the following tools, for which we also give an example workflow in Appendix 4 and documentation at https://idtracker.ai/latest/user_guide/tools.html:

idtrackerai_inspect_clusters, a tool to visually inspect the learned representation space and check the clusters integrity.

idtrackerai_validate, a graphic app to review and correct tracking results.

idtrackerai_video, a graphic app to generate videos of the computed animal trajectories on top of the original video for visualization. This app also generates individual videos for each animal showing only its cropped region over time to be able to run pose estimators like the ones in Lauer et al. (2022); Pereira et al. (2022); Segalin et al. (2021); Tang et al. (2025); Biderman et al. (2024).

idmatcher.ai, a tool to match identities across multiple recordings (see Appendix 5).

Direct integration with SocialNet, a model of collective behavior introduced in Heras et al. (2019).

Direct integration with trajectorytools, a Python package for 2D trajectory processing, and a set of Jupyter Notebooks that uses trajectorytools to analyze basic movement properties and spatial relationships between the animals.

Methods

Software availability

id-tracker.ai is a free and open source project (license GPLv3). Information about its installation and usage can be found on the website https://idtracker.ai/. The source code is available in https://gitlab.com/polavieja_lab/idtrackerai and the package is pip-installable from PyPI. All versions can be found in these platforms, specifically “original idtracker.ai (v4)” as v4.0.12, “optimized v4 (v5)” as v5.2.12 and “new idtracker.ai (v6)” as v6.0.9. We only actively maintain and provide support for the latest version available, having the old ones for archive and reference only.

Tested computer specifications

The software idtracker.ai depends on PyTorch and is thus compatible with any machine that can run PyTorch, including Windows, MacOS, and Linux systems. Although no specific hardware is required, a graphics card is highly recommended for hardware-accelerated machine-learning computations.

Version 6 of idtracker.ai was tested on computers running Ubuntu 24.04, Fedora 41, Windows 11 with NVIDIA GPUs from the 1000 to the 4000 series, and MacOS 15 with Metal chips. The benchmark results presented in this study were performed on a desktop computer running Ubuntu 24.04 LTS 64bit with a AMD Ryzen 9 5950X (32 cores at 3.4 GHz) processor, 128 GB RAM and an NVIDIA GeForce RTX 4090.

Occlusion tests

We clarify here some details from the occlusion tests presented in the section 'No need for global fragments'. The occlusion tests were performed using the same set of videos as in the benchmark. For these tests, we defined an occlusion mask as a region of interest in the software, applied during the segmentation step. When video frames were converted into binary foreground-background images, all pixels inside the mask were treated as background.

The fragments are detected as in any other video. Specifically, this means that when one animal is hidden in the mask, the animal is lost and the fragment breaks. Fragments are not built adding artificial links.

For evaluation, the ground truth contained the positions of all animals at all times, but, obviously, only positions outside the mask were used to compute tracking accuracy. To avoid partial detections of the animals, trajectories within 15 pixels of the mask boundary (75 pixels for mice videos) were excluded from the evaluation.

eLife Assessment

This important study introduces an advance in multi-animal tracking by reframing identity assignment as a self-supervised contrastive representation learning problem. It eliminates the need for segments of video where all animals are simultaneously visible and individually identifiable, and significantly improves tracking speed, accuracy, and robustness with respect to occlusion. This innovation has implications beyond animal tracking, potentially connecting with advances in behavioral analysis and computer vision. The strength of support for these advances is compelling overall, although there were some remaining minor methodological concerns.

Reviewer #1 (Public review):

Summary:

This is a strong paper that presents a clear advance in multi-animal tracking. The authors introduce an updated version of idtracker.ai that reframes identity assignment as a contrastive representation learning problem rather than a classification task requiring global fragments. This change leads to substantial gains in speed and accuracy and removes a known bottleneck in the original system. The benchmarking across species is comprehensive, the results are convincing, and the work significant.

Strengths:

The main strengths are the conceptual shift from classification to representation learning, the clear performance gains, and the improved robustness of the new version. Removing the need for global fragments makes the software much more flexible in practice, and the accuracy and speed improvements are well demonstrated across a diverse set of datasets. The authors' response also provides further support for the method's robustness.

The comparison to other methods is now better documented. The authors clarify which features are used, how failures are defined, how parameters are sampled, and how accuracy is assessed against human-validated data. This helps ensure that the evaluation is fair and that readers can understand the assumptions behind the benchmarks.

The software appears thoughtfully implemented, with GUI updates, integration with pose estimators, and tools such as idmatcher.ai for linking identities across videos. The overall presentation has been improved so that the limitations of the original idtracker.ai, the engineering optimizations, and the new contrastive formulation are more clearly separated. This makes the central ideas and contributions easier to follow.

Weaknesses:

I do not have major remaining criticisms. The authors have addressed my earlier concerns about the clarity and fairness of the comparison with prior methods, the benchmark design, and the memory usage analysis by adding methodological detail and clearly explaining their choices. At this point I view these aspects as transparent features of the experimental design that readers can take into account, rather than weaknesses of the work.

Overall, this is a high-quality paper. The improvements to idtracker.ai are well justified and practically significant, and the authors' response addresses the main concerns about clarity and evaluation. The conceptual contribution, thorough empirical validation, and thoughtful software implementation make this a valuable and impactful contribution to multi-animal tracking.

Reviewer #3 (Public review):

Summary:

The authors propose a new version of idTracker.ai for animal tracking. Specifically, they apply contrastive learning to embed cropped images of animals into a feature space where clusters correspond to individual animal identities. By doing this, they address the requirement for so-called global fragments - segments of the video, in which all entities are visible/detected at the same time. In general, the new method reduces the long tracking times from the previous versions, while also increasing the average accuracy of assigning the identity labels.

Strengths and weaknesses:

The authors have reorganized and rewritten a substantial portion of their manuscript, which has improved the overall clarity and structure to some extent. In particular, omitting the different protocols enhanced readability. However, all technical details are now in appendix which is now referred to more frequently in the manuscript, which was already the case in the initial submission. These frequent references to the appendix - and even to appendices from previous versions - make it difficult to read and fully understand the method and the evaluations in detail. A more self-contained description of the method within the main text would be highly appreciated.

Furthermore, the authors state that they changed their evaluation metric from accuracy to IDF1. However, throughout the manuscript they continue to refer to "accuracy" when evaluating and comparing results. It is unclear which accuracy metric was used or whether the authors are confusing the two metrics. This point needs clarification, as IDF1 is not an "accuracy" measure but rather an F1-score over identity assignments.

The authors compare the speedups of the new version with those of the previous ones by taking the average. However, it appears that there are striking outliers in the tracking performance data (see Supplementary Table 1-4). Therefore, using the average may not be the most appropriate way to compare. The authors should consider using the median or providing more detailed statistics (e.g., boxplots) to better illustrate the distributions.

The authors did not provide any conclusion or discussion section. Including a concise conclusion that summarizes the main findings and their implications would help to convey the message of the manuscript.

The authors report an improvement in the mean accuracy across all benchmarks from 99.49% to 99.82% (with crossings). While this represents a slight improvement, the datasets used for benchmarking seem relatively simple and already largely "solved". Therefore, the impact of this work on the field may be limited. It would be more informative to evaluate the method on more challenging datasets that include frequent occlusions, crossings, or animals with similar appearances. The accuracy reported in the main text is "without crossings" - this seems like incomplete evaluation, especially that tracking objects that do not cross seems a straightforward task. Information is missing why crossings are a problem and are dealt with separately. There are several videos with a much lower tracking accuracy, explaining what the challenges of these videos are and why the method fails in such cases would help to understand the method's usability and weak points.

Author response:

The following is the authors’ response to the previous reviews

eLife Assessment

This important study introduces an advance in multi-animal tracking by reframing identity assignment as a self-supervised contrastive representation learning problem. It eliminates the need for segments of video where all animals are simultaneously visible and individually identifiable, and significantly improves tracking speed, accuracy, and robustness with respect to occlusion. This innovation has implications beyond animal tracking, potentially connecting with advances in behavioral analysis and computer vision. The strength of support for these advances is compelling overall, although there were some remaining minor methodological concerns.

To tackle “minor methodological concerns” mentioned in the Editorial assessment and Reviewer 3, the new version of the manuscript includes the following changes:

a) The new ms does not anymore use the word “accuracy” but “IDF1 scores”. See, for example, Lines 46, 161, 176, and 522 for our new wording as “IDF1 scores”.

b) Instead of comparing softwares using mean accuracy over the benchmark, Reviewer 3 proposes to use medians or even boxplots. We now provide boxplot results with mean, median, percentiles and outliers (Figure 1- figure Supplement 2).

Additionally, we also include in the text the other recommendations from Reviewer 3:

a) We now more explicitly describe the problems of the original idtracker.ai v4 in the benchmark (lines 66-68). Around half of the videos had a high accuracy in the original dtracker.ai (v4) but the other half of the videos had lower accuracies (Figure 1a, blue). The new idtracker.ai has high accuracy values for all the videos (Figure 1a, magenta).

Also, the videos with high accuracy in the old idtracker.ai had very long tracking times (Figure 1b, blue) and the new version does not (Figure 1b, magenta). So the benchmark allows us to distinguish the new idtracker.ai as having a better accuracy for all videos and lower tracking times, making it a much more practical system than previous ones.

b) We further clarified the occlusion experiment (lines 188-190 and 277-290).

c) We explain why we measure accuracies with and without animal crossings (lines 49-62).

d) We added a Discussion section (lines 223-244).

We believe the new version has clarified the minor methodological concerns.

Reviewer #3 (Public review):

The authors have reorganized and rewritten a substantial portion of their manuscript, which has improved the overall clarity and structure to some extent. In particular, omitting the different protocols enhanced readability. However, all technical details are now in appendix which is now referred to more frequently in the manuscript, which was already the case in the initial submission. These frequent references to the appendix - and even to appendices from previous versions - make it difficult to read and fully understand the method and the evaluations in detail. A more self-contained description of the method within the main text would be highly appreciated.

In the new ms, we have reduced the references to the appendix by having a more detailed explanation in one place, lines 49-62.

Furthermore, the authors state that they changed their evaluation metric from accuracy to IDF1. However, throughout the manuscript they continue to refer to "accuracy" when evaluating and comparing results. It is unclear which accuracy metric was used or whether the authors are confusing the two metrics. This point needs clarification, as IDF1 is not an "accuracy" measure but rather an F1-score over identity assignments.

We thank the reviewer for noticing this. Following this recommendation, we changed how we refer to the accuracy measure with “IDF1 score” in the entire ms. See, for example, lines 46, 161, 176, and 522.

The authors compare the speedups of the new version with those of the previous ones by taking the average. However, it appears that there are striking outliers in the tracking performance data (see Supplementary Table 1-4). Therefore, using the average may not be the most appropriate way to compare. The authors should consider using the median or providing more detailed statistics (e.g., boxplots) to better illustrate the distributions.

We thank the reviewer for asking for more detailed statistics. We added the requested box plot in Figure 1- figure Supplement 2 to provide more complete statistics in the comparison.

The authors did not provide any conclusion or discussion section. Including a concise conclusion that summarizes the main findings and their implications would help to convey the message of the manuscript.

We added a Discussion section in lines 223-244.

The authors report an improvement in the mean accuracy across all benchmarks from 99.49% to 99.82% (with crossings). While this represents a slight improvement, the datasets used for benchmarking seem relatively simple and already largely "solved". Therefore, the impact of this work on the field may be limited. It would be more informative to evaluate the method on more challenging datasets that include frequent occlusions, crossings, or animals with similar appearances.

Around half of the videos also had a very high accuracy in the original dtracker.ai (v4) but the other half of the videos had lower accuracies (Figure 1a, blue). For example, we found IDF1 scores of 94.47% for a video of 100 zebrafish with thousands of crossings (z_100_1), 93.77% for a video of 4 mice (m_4_2) and 69.66% for a video of 100 flies (d_100_3). The new idtracker.ai has high accuracy values for all the videos (Figure 1a, magenta).

Importantly, the tracking times for the majority of videos was very high in the original idtracker.ai (Figure 1b, blue), making the use of the tracking system limited in practice. The new system manages both a high accuracy in all videos (Figure 1a, magenta) and much lower tracking times (Figure 1b, magenta), making it a much more practical system..

We have added a sentence of the limitations of the original idtracker.ai as obtained from the benchmark, lines 66-68.

The accuracy reported in the main text is "without crossings" - this seems like incomplete evaluation, especially that tracking objects that do not cross seems a straightforward task. Information is missing why crossings are a problem and are dealt with separately.

We have now added an explanation on why we measure accuracy without crossings and why we separated it from the accuracy for all the trajectory in lines 49-62. The reason is that the identification algorithm being presented in this ms only identifies animal images outside the crossings. This algorithm makes robust animal identifications through the video despite the thousands of animal crossings typically existing in each of our videos used in the benchmark. It is a second algorithm (that hasn’t changed since the first idTracker in 2014) the one that assigns animal positions during crossings once the first algorithm has made animal identifications before and after the crossings.

There are several videos with a much lower tracking accuracy, explaining what the challenges of these videos are and why the method fails in such cases would help to understand the method's usability and weak points.

Some videos had low accuracy on previous versions (Figure 1a, blue), but the new idtracker.ai has high accuracy in all of them (Figure 1a, magenta).

Reviewer #3 (Recommendations for the authors):

(1) As described before, the authors claim to use IDF1 as their metric in the whole manuscript (lines 414-436) but only refer to accuracy when presenting the results. It is not clear, whether accuracy was used as a metric instead of IDF1 or the authors are confusing these metrics.

Following this recommendation, we replaced “accuracy” with “IDF1 score” , see lines 46, 161, 176, and 522.

(2) In the introduction, a brief explanation why crossings need to be dealt with separately would help to understand the logic of the method design.

We added such an explanation in lines 49-62.

(3) Figure 3: We asked about how the tracking accuracy is being assessed with occlusions. The authors responded with that only the GT points inside the ROI are taken into account when computing the accuracy. Does this mean, that the occluded blobs are still part of the CNN training and the clustering? This questions the purpose of this experiment, since the accuracy performance would therefore only change, if the errors, that their approach is doing either way, are outside the ROI and, therefore, not part of the metric evaluation.

The occluded blobs are not part of any training because they are erased from the video, they do not exist. We made this more clear in lines 188-190 and 277-290.

(4) Figure 1: The fact that datasets are connected with a line is misleading - there is no connection between the data along the x-axis. A line plot is not an appropriate way to present these results.

The new ms clarifies that the lines are for ease of visualization, see last line in the caption of Figure 1.

(5) Lines 38-39: It is not clear how the CNN can be pretrained for the entire video if there are no global segments or only short ones. Here, the distinction between "no segments", "only short segments" and "pretraining on the entire video" is not explained.

This pretraining protocol is not used in the version of the software we present, so details of this are not as relevant.

(6) Figure 2a: The authors are showing "individual fragments" and individual fragments in a global fragment." However, it seems there are a few blue borders missing. In the text (l. 73-79), they note, that they are displaying them as "examples" but the absence of correct blue borders is confusing.

In the new ms, we have replaced the label “Individual fragments in a global fragment” with “Individual fragments in an example global fragment” in the legend of Figure 2.

(7) Lines 61-63, 148-151, and 162-164: Could the authors clarify why they used the average instead of median when comparing the speedups of the new version and the old ones?

We thank the reviewer for asking for more detailed statistics. We added the requested box plot in Figure 1- figure Supplement 2 to provide more complete statistics in the comparison of accuracies and tracking times for old and new systems.

(8) Lines 140-144: The post-processing steps are not clear. The authors should rather state clearly which processes of the old versions they are using. Then the authors could shortly explain them.

We removed this paragraph and explained in more detail in lines 49-62 which parts of the software are new and which ones are not.

(9) Lines 239-251: Here, the authors are clarifying on a section 1-2 pages before. This information should be directly in that section instead.

Following this recommendation, we clarified the occlusion experiment in the main text (lines 188-191) to make it more self-contained. Still, the flow of the main text is better with some details in Methods.

(10) Line 38: It is not clear how the CNN can be pretrained for the entire video if there are no global segments or only short ones. Here, the distinction between "no segments"

"only short segments" and "pretraining on the entire video" is a bit misleading/underexplained.

See number 5.

(11) Figure 2a: The authors are showing "individual fragments" and individual fragments in a global fragment." However, it seems there are a few blue borders missing. In the text (l. 73-79), they note, that they are displaying them as "examples" but the absence of correct blue borders is confusing.

See number 6.

(12) Figure 2c and line 115-118: "Batches" itself is not meaningful without any information of the batch size. The authors should rather depict the batch size and then the number of epochs. The Figure 2 contains the info 400 positive and 400 negative pairs of images per batch. However, there is no information about the total number of images.

Furthermore, these metrics are inappropriate here, since training is carried out from scratch (or already pre-trained) for every new video, each video has different number of animals, different number of images.

Following this recommendation, we clarified the number of images in each batch (Figure 1c caption and lines 134-138), why we do not work with epochs (lines 700-702), and the idea that the clusters in Figure 2 represent an example and the number of batches needed for the clusters to form depends on the video details.

Appendix 1-figure 1: why do the methods fail? It looks that for certain videos the method is fairly unreliable. What is the reason for the methods to crash and how to avoid this?

Those failures are only for the old idtracker.ai and Trex, not for the method presented here. Our new contrastive algorithm does not fail in any of the videos in the benchmark.

We thank the reviewer for the detailed suggestions. We believe we have incorporated all of them in the new version of the ms.

Version 3

Abstract

idTracker and idtracker.ai approach multi-animal tracking from video as an image classification problem. For this classification, both rely on segments of video where all animals are visible to extract images and their identity labels. When these segments are too short, tracking can become slow and inaccurate and, if they are absent, tracking is impossible. Here, we introduce a new idtracker.ai that reframes multi-animal tracking as a representation learning problem rather than a classification task. Specifically, we apply contrastive learning to image pairs that, based on video structure, are known to belong to the same or different identities. This approach maps animal images into a representation space where they cluster by animal identity. As a result, the new idtracker.ai eliminates the need for video segments with all animals visible, is more accurate, and tracks up to 700 times faster.

Main text

Video-tracking systems that attempt to follow individuals frame-by-frame can fail during occlusions, resulting in identity swaps that accumulate over time Branson et al. (2009); Plum (2024); Chen et al. (2023); Chiara and Kim (2023); Liu et al. (2023); Bernardes et al. (2021). idTracker Pérez-Escudero et al. (2014) introduced the paradigm of animal tracking by identification from the animal images. This approach, unfeasible for humans, avoids the accumulation of errors by identity swaps during occlusions. Its successor, idtracker.ai Romero-Ferrero et al. (2019), built on this paradigm by incorporating deep learning and achieved accuracies often exceeding 99.9% in videos of up to 100 animals.

Limitation of the original idtracker.ai

The core idea of the original idtracker.ai is to use a segment of the video in which all animals are visible to extract a set of images and identity labels for each individual. These labeled images are used to train a convolutional neural network (CNN) with one class per animal. Once trained, the network assigns identities to other segments in which all animals are visible. Only segments with identity assignments that meet strict quality criteria are retained, and their images and labels are also used for further training of the CNN. This iterative process of training, assigning, and selecting continues until most of the animal images in the video have been assigned to identities.

If no segment exists in which all animals are visible, the original idtracker.ai cannot start. More commonly, such segments can be short and result in a CNN of low quality. To improve performance in this case, the original idtracker.ai pretrains the CNN using the entire video, but this process is slow. As a consequence, when the segments in which all animals are visible are too short, accuracy might be lower and tracking time longer.

We built a benchmark to test this old version of idtracker.ai against the new ones. We quantified tracking accuracy using the standard Identification F1 Score (IDF1) (see Appendix 1). We rely on IDF1 because it reflects whether trajectories preserve correct identity assignments throughout the video. Figure 1a (blue line) gives the median IDF1 scores for the original idtracker.ai in our benchmark of 33 videos. The first 15 videos of the benchmark are videos of zebrafish, flies (drosophila) and mice for which of the original idtracker.ai has IDF1 score of > 99.9%. In the remaining videos the IDF1 score decreases, reaching 93.77% in video 𝑚_4_2, and 69.66% in video 𝑑_100_3, which lies outside the plotted range.

Performance for a benchmark of 33 videos of flies, zebrafish and mice.

a. Median IDF1 score computed using all images of animals in the videos excluding animal crossings. The videos are ordered by decreasing IDF1 score of the original idtracker.ai results for ease of visualization. b. Median tracking times are shown for the scale of hours and, in the inset, for the scale of days. The videos are ordered by increasing tracking times in the original idtracker.ai results for ease of visualization. Supplementary Table 1, Supplementary Table 2, Supplementary Table 3 give more complete statistics (median, mean and 20-80 percentiles) for the original idtracker.ai (version 4 of the software), optimized v4 (version 5) and new idtracker.ai (version 6), respectively. The names of the videos in a. and b. start with a letter for the species (𝑧,𝑑,𝑚), followed by the number of animals in the video, and possibly an extra number to distinguish the video if there are several of the same species and animal group size. Lines between points are for visualization purposes only. Figure 1—figure supplement 1. Performance for the benchmark with full trajectories with animal crossings Figure 1—figure supplement 2. Boxplot representation of the benchmark results Figure 1—figure supplement 3. Robustness to blurring and light conditions Figure 1—figure supplement 4. Memory usage across the different softwares

These accuracies correspond to all animal images in the video excluding images where animals cross paths. We isolate this metric because the core identification algorithm, which is the primary novelty presented in this article, works on these single-animal images to track identities despite the thousands of crossings typically existing in each video. Following this identification, idtracker.ai applies a deterministic post-processing pipeline. This pipeline includes the correction of misidentifications by enforcing the same identity to all individual images between two animal crossings and by detecting impossible abrupt changes in location. It also includes computing animal positions during crossings using the predicted identities from immediately before and after the crossing. This post-processing pipeline remains unchanged from the original idTracker publication Pérez-Escudero et al. (2014), where its details are fully described. We also report the accuracy as IDF1 scores for the complete video including crossings (Figure 1—figure Supplement 1) as this is the final accuracy the user experiences. The data used for figures Figure 1 and Figure 1—figure Supplement 1 can be found in Supplementary Table 1, and further details regarding the benchmark are available in Appendix 1.

Figure 1b (blue line) shows the median times that the original idtracker.ai takes to track each of the videos in the benchmark. Some of the videos take a few minutes to track, others a few hours, and six videos take more than three days, one nearly two weeks.

We learn from the benchmark of the original idtracker.ai that it fails to accurately track animals in challenging videos and its computational time can be of days or weeks, in practice a bottleneck in the study of group behavior from video.

Optimizing idtracker.ai without changes in the learning method

We first optimized idtracker.ai without changing how we identify animals. We improved data loading and redesigned the main objects in the software (see Appendix 2 for details). This version of the optimized original idtracker.ai (version 5 of the software) achieved higher accuracies, Figure 1a (orange line), and Figure 1—figure Supplement 1a (orange line) for results that include animal crossings (data for these tests can be found at Supplementary Table 2). The mean IDF1 score across the benchmark is 99.63% without crossings and 99.49% with crossings, compared with 98.39% without crossings and 98.24% with crossings for the original idtracker.ai (see Figure 1—figure Supplement 2a and b for boxplots showing more statistics).

Tracking times were also significantly reduced. As shown in Figure 1b (orange line), no video took longer than a day. On average, tracking is 13.5 times faster than with the original version and 120.1 times faster for the more difficult videos (see Figure 1—figure Supplement 2c for boxplots showing more statistics). Despite the large improvement in tracking times, some of them are many hours or close to a day, which would still be too limiting in many pipelines.

To further improve both accuracy and tracking speed, we retained these optimizations while changing the core identification algorithm to eliminate the need for segments of video in which all animals are visible.

The new idtracker.ai uses representation learning

We reformulate multi-animal tracking as a representation learning problem. In representation learning, we learn a transformation of the input data that makes it easier to perform downstream tasks Xing et al. (2002); Bengio et al. (2013); Ericsson et al. (2022). In our case, the downstream task is to cluster images into animal identities without requiring identity labels.

This reformulation is made possible by the inherent structure of the video, illustrated in Figure 2a. First, we detect specific moments in the video when animals touch or cross paths. These are shown in Figure 2a as boxes with dashed borders that contain images of overlapping animals. We can then divide the rest of the video into individual fragments, each consisting of the set of images of a single individual between two animal crossings. Figure 2a shows 14 such fragments as rectangles with a gray background. In addition, a video with 𝑁animals may contain global fragments, that is, a collection of 𝑁 individual fragments that coexist in one or more consecutive frames. An example is shown in Figure 2a by the five fragments with blue borders. These global fragments are used by the original idtracker.ai to train a CNN. In the new approach, we do not assume that global fragments exist.

Tracking by identification using deep contrastive learning.

a Schematic representation of a video with five fish. b A ResNet18 network with 8 outputs generates a representation of each animal image as a point in an 8-dimensional space (here shown in 2D for visualization). Each pair of images corresponds to two points in this space, separated by a Euclidean distance. The ResNet18 network is trained to minimize this distance for positive pairs and maximize it for negative pairs. c 2D t-SNE visualizations of the learned 8-dimensional representation space. Each dot represents an image of an animal from the video. As training progresses, clusters corresponding to individual animals become clearer. Here we plot this process for the example video zebrafish_60_1 after training for 0, 2,000, 4,000 and 15,000 batches (each batch contains 400 positive and 400 negative pairs of images, that is, 1,600 images per batch). The t-SNE plot at 15,000 training batches is also shown color-coded by human-validated ground-truth identities. The pink rectangle at 2,000 batches of training highlights clear clusters and the orange square fuzzy clusters. d The Silhouette score measures cluster coherence and increases during training, as illustrated for a video with 60 zebrafish. e A Silhouette score of 0.91 corresponds to a human-validated error rate of less than 1% per image.

Representation learning becomes feasible in this context because we can obtain both positive and negative image pairs without using identity labels. Positive pairs are images of the same individual obtained from within the same individual fragment (Figure 2a, green boxes). Negative pairs are images of different individuals taken from different individual fragments that coexist in time for one or more frames (Figure 2a, red boxes). We can then use the positive and negative pairs of images for contrastive learning, a self-supervised learning framework designed to learn a representation space in which positive examples are close together, and negative examples are far apart Schroff et al. (2015); Dong and Shen (2018); KAYA and BİLGE (2019); Chen et al. (2020a,b); Guo et al. (2020); Wang et al. (2020); Yang et al. (2020). This formulation is supported by the success of deep metric and contrastive learning methods (see Appendix 3 for comparison with previous work).

We first evaluated which neural networks are suitable for contrastive learning with animal images. In addition to our previous CNN from idtracker.ai, we tested 31 networks from 10 different families of state-of-the-art convolutional neural networks and transformer architectures, selected for their compatibility with consumer-grade GPUs and ability to handle small input images (20 × 20 to 100 × 100 pixels) typical in collective animal behavior videos. Among these architectures, ResNet18 (v1) He et al. (2016a) without pretrained weights was the fastest to obtain low errors (see Appendix 3).

A ResNet18 with 𝑀 outputs maps each input image to a point in an 𝑀-dimensional representation space (illustrated in Figure 2b as a point on a plane). Experiments showed that using 𝑀 = 8 achieved faster convergence to low error (see Appendix 3). ResNet18 is trained using a contrastive loss function (Chopra et al. (2005), see Appendix 3 for details). Each image in a positive or negative pair is input separately into the network, producing a point in the 8-dimensional representation space. For an image pair, we then obtain two points in an 8-dimensional space, separated by some (Euclidean) distance. The optimization of the loss function minimizes (or maximizes) this Euclidean distance for positive (or negative) pairs until the distance 𝐷pos (or 𝐷neg) is reached. The effect of 𝐷pos is to prevent the collapse to a single of the positive images coming from the same fragment, allowing for a small region of the 8-dimensional representation space to contain all the positive pairs of the same identity. The effect of 𝐷neg is to prevent excessive scatter of the points representing images from negative pairs. We empirically determined that 𝐷neg∕𝐷pos = 10 results in a faster method to obtain low error (see Appendix 3), and we use 𝐷pos = 1 and 𝐷neg = 10.

As the model trains, the representation space becomes increasingly structured, with similar data points forming coherent clusters. Figure 2c visualizes this progression using 2D t-SNE van der Maaten and Hinton (2008) plots of the 8-dimensional representation space. After 2, 000 training batches (400 positive and 400 negative pairs of images per batch), initial clusters emerge, and by 15,000 batches, distinct clusters corresponding to individual animals are evident. Ground truth identities verified by humans confirm that each cluster corresponds to an animal identity (Figure 2c, colored clusters).

The method to select positive and negative pairs is critical for fast learning Awasthi et al. (2022); Khosla et al. (2021); Rösch et al. (2024). This is because not all image pairs contribute equally to training. Figure 2c shows at 2, 000 training batches that some clusters are well-defined (e.g. those inside the pink rectangle) while others remain fuzzy (e.g. those inside the orange square). Images in well-defined clusters have negligible impact on the loss or weight updates, as positive pairs are already close and negative pairs are sufficiently separated. Sampling from these well-defined clusters, therefore, wastes time. In contrast, fuzzy clusters contain images that still contribute significantly to the loss and benefit from further training. To address this, we developed a sampling method that prioritizes pairs from underperforming clusters requiring additional learning, while maintaining baseline sampling for all clusters based on fragment size (see Appendix 3). This ensures consistent updates across the representation space and prevents forgetting in well-defined clusters.

To assign identities to animal images, we perform K-means clustering Sculley (2010) on the points representing all images of the video in the learned 8-dimensional representation space. Each image is then assigned to a cluster with a probability that increases the closer it is to the cluster center. To evaluate clustering quality, we compute the mean Silhouette score Rousseeuw (1987), which quantifies intra-cluster cohesion and inter-cluster separation. A maximum value of 1 indicates ideal clustering. During training, the mean Silhouette score increases (Figure 2d). We empirically determined that a value of 0.91 for this index corresponds to an identity assignment error below 1% for a single image (Figure 2e). As a result, we use 0.91 as the stopping criterion for training (see Appendix 3).

The new idtracker.ai (v6) is more accurate than original idtracker.ai (v4) and than its optimized version (v5), Figure 1a (purple line), with data at Supplementary Table 3. Its average IDF1 score in the benchmark is 99.92% and 99.82% without and with crossings, respectively, an important improvement over the original idtracker.ai v4 (98.39% and 98.24%) and its optimized version v5 (99.63% and 99.49%). It also gives much shorter times than the original idtracker.ai (v4) and its optimized version (v5), Figure 1b (purple line). It is on average 90 times faster than the original idtracker.ai (v4) and, for the more difficult videos, up to 712 times faster. See Figure 1—figure Supplement 2 for boxplots showing more statistics comparing tracking systems.

As for the original idtracker.ai, the new idtracker.ai can work well with lower resolutions, blur and video compression, and with inhomogeneous light (Figure 1—figure Supplement 3). We also compared the new idtracker.ai to TRex Walter and Couzin (2021), which is based on idtracker.ai without pretraining and with additional operations like eroding crossings to make global fragments longer, posture image normalization, tracklet subsampling, and the use of uniqueness feedback during training. TRex gives comparable accuracies to the original idtracker.ai in the benchmark, and it is on average 31 times faster than the original idtracker.ai and up to 316 times faster (Figure 1—figure Supplement 1b, with data at Supplementary Table 4). However, the new idtracker.ai is both more accurate and faster than TRex (Figure 1—figure Supplement 1). The mean IDF1 score of TRex across the benchmark is 98.14% and 97.89% excluding and including animal crossings, respectively. This is noticeably below the values for the new idtracker.ai of 99.92% and 99.82%, respectively. Also, the new idtracker.ai is on average 4.5 times faster and up to 16.5 times faster than TRex. See Figure 1—figure Supplement 2 for boxplots showing more statistics for IDF1 scores and tracking times. Additionally, the new idtracker.ai has a memory peak lower than TRex (Figure 1—figure Supplement 4).

No need for global fragments

The new idtracker.ai also works in videos in which the original idtracker.ai does not even track because there are no global fragments. Global fragments are absent in videos with very extensive animal occlusions, for example because animals touch or cross more frequently, parts of the setup are covered, or the camera focuses on only a specific region of the setup.

To systematically evaluate this, we tracked videos in which we erased sectors of angle 𝜃(Figure 3). These sectors are painted black before the tracking starts so animals inside this region are not visible to the tracking systems (see Methods for more details about the occlusion tests).

The light and dark gray regions in Figure 3 correspond to videos with no global fragments, and the original idtracker.ai and its optimized version declare tracking impossible in these regions. The new idtracker.ai, however, works well until approximately 1∕4 of the setup is visible, and afterward it degrades. This shows the limit of the new idtracker.ai. The deterioration happens because, for the clustering process to be successful, we need enough coexisting individual fragments to have both positive and negative examples. We measure this using fragment connectivity, defined as the average number of other fragments a fragment coexists with, divided by 𝑁− 1, with 𝑁 the total number of animals in the video. Empirically, we find that a fragment connectivity below 0.5 corresponds to low accuracies Figure 3—figure Supplement 1. The new idtracker.ai warns the user when this condition of low fragment connectivity takes place, which we indicate in Figure 3 with the dark gray regions.

Tracking with strong occlusions.

Accuracies when we mask a region of a video defined by an angle 𝜃 and the tracking system has no access to the information behind the mask. Light and dark gray region correspond to the angles for which no global fragments exist in the video. Dark gray regions correspond to angles for which the video has a fragment connectivity lower than 0.5, with the fragment connectivity defined as the average number of other fragments each fragment coexists with, divided by 𝑁− 1, with 𝑁 the total number of animals; see Figure 3—figure Supplement 1, for an analysis justifying this value of 0.5. The original idtracker.ai (v4) and its optimized version (v5) cannot work in the gray regions, and new idtracker.ai is expected to deteriorate only in the dark gray region. Figure 3—figure supplement 1. Fragment connectivity analysis

Output of new idtracker.ai

The final output of the new idtracker.ai consists of the 𝑥 and 𝑦 coordinates for each identified animal and video frame. Additionally, it provides an estimation of the achieved accuracy, the Silhouette score as a measure of clustering quality, and the probability of correct identity assignment for each animal and frame. The new idtracker.ai also includes the following tools, for which we also give an example workflow in Appendix 4 and documentation at https://idtracker.ai/latest/user_guide/tools.html:

idtrackerai_inspect_clusters, a tool to visually inspect the learned representation space and check the clusters integrity.

idtrackerai_validate, a graphic app to review and correct tracking results.

idtrackerai_video, a graphic app to generate videos of the computed animal trajectories on top of the original video for visualization. This app also generates individual videos for each animal showing only its cropped region over time to be able to run pose estimators like the ones in Lauer et al. (2022); Pereira et al. (2022); Segalin et al. (2021); Tang et al. (2025); Biderman et al. (2024).

idmatcher.ai, a tool to match identities across multiple recordings (see Appendix 5).

Direct integration with SocialNet, a model of collective behavior introduced in Heras et al. (2019).

Direct integration with trajectorytools, a Python package for 2D trajectory processing, and a set of Jupyter Notebooks that uses trajectorytools to analyze basic movement properties and spatial relationships between the animals.

Discussion

We have introduced a new way to perform multi-animal tracking that shifts from a classification task to a representation learning one. By leveraging contrastive learning on image pairs derived from the temporal structure of the video, the new idtracker.ai eliminates the restrictive requirement for segments where all animals are simultaneously visible. The idea of contrastive identification could also be of value in other contexts, for example in tracking body parts.

The new idtracker.ai is not only more robust to occlusions but also outperforms previous tracking systems both in accuracy and tracking time. With a median IDF1 score of 99.92% and processing speeds up to 700 times faster than previous versions, the software transforms tracking from a computational bottleneck requiring days or weeks of processing into a much faster step suitable for agile experimental loops.

This approach expands the scope of feasible behavioral studies, allowing for more complex environments where animals are frequently occluded or move in and out of the field of view. Our experiments with masked video regions demonstrate that the system requires only a minimum sufficient pairwise co-occurrence of animals rather than complete group visibility, making it robust to conditions where previous algorithms fail.

Additionally, we have surrounded this new core algorithm with a software ecosystem to improve user experience. We introduced interactive tools that allow researchers to visually inspect the learned clusters and easily validate trajectories. We also prioritized interoperability by facilitating the generation of data for pose-estimation frameworks like DeepLabCut Lauer et al. (2022) and SLEAP Pereira et al. (2022) for fine-grained body analysis, and the direct integration with modeling of group behavior like SocialNet Heras et al. (2019).

Methods

Software availability

idtracker.ai is a free and open source project (license GPLv3). Information about its installation and usage can be found on the website https://idtracker.ai/. The source code is available in https://gitlab.com/polavieja_lab/idtrackerai and the package is pip-installable from PyPI. All versions can be found in these platforms, specifically “original idtracker.ai (v4)” as v4.0.12, “optimized v4 (v5)” as v5.2.12 and “new idtracker.ai (v6)” as v6.0.9. We only actively maintain and provide support for the latest version available, having the old ones for archive and reference only.

Tested computer specifications

The software idtracker.ai depends on PyTorch and is thus compatible with any machine that can run PyTorch, including Windows, MacOS, and Linux systems. Although no specific hardware is required, a graphics card is highly recommended for hardware-accelerated machine-learning computations.

Version 6 of idtracker.ai was tested on computers running Ubuntu 24.04, Fedora 41, Windows 11 with NVIDIA GPUs from the 1000 to the 4000 series, and MacOS 15 with Metal chips. The benchmark results presented in this study were performed on a desktop computer running Ubuntu 24.04 LTS 64bit with a AMD Ryzen 9 5950X (32 cores at 3.4 GHz) processor, 128 GB RAM and an NVIDIA GeForce RTX 4090.

Occlusion tests

We clarify here some details from the occlusion tests presented in the section No need for global fragments. The occlusion tests were performed using the same set of videos as in the benchmark. For these tests, we defined an occlusion mask as a region of interest in the software. When video frames are converted into binary foreground-background images, all pixels inside the mask are treated as background so no information is extracted from them.

The fragments are detected as in any other video. Specifically, this means that when one animal is hidden in the mask, the animal is lost and the fragment breaks. No hidden visual information outside the region of interest is used to track the visible portion so fragments are not built adding any artificial links.

For evaluation, the ground truth of the unmasked video containing the positions of all animals at all times was used, but, obviously, only positions outside the mask were used to compute tracking accuracy. To avoid partial detections of the animals, trajectories within 15 pixels of the mask boundary (75 pixels for mice videos) were excluded from the evaluation.